---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
has_callme <- requireNamespace("callme", quietly = TRUE)
has_bench <- requireNamespace("bench", quietly = TRUE)
run_ssh_demos <- tolower(Sys.getenv("RTINYCC_RUN_SSH_DEMOS")) %in%
  c("1", "true", "yes")
```

# Rtinycc

 Builds `TinyCC` `Cli` and Library For `C` Scripting in `R`

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rtinycc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rtinycc/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/Rtinycc)](https://CRAN.R-project.org/package=Rtinycc)
[![Rtinycc status badge](https://sounkou-bioinfo.r-universe.dev/Rtinycc/badges/version)](https://sounkou-bioinfo.r-universe.dev/Rtinycc)
<!-- badges: end -->

## Abstract

Rtinycc is an R interface to [TinyCC](https://github.com/TinyCC/tinycc), providing both CLI access and a libtcc-backed in-memory compiler. It includes an FFI inspired by [Bun's FFI](https://bun.com/docs/runtime/ffi) for binding C symbols with predictable type conversions and pointer utilities. The package works on unix-alikes and Windows and focuses on embedding TinyCC and enabling JIT-compiled bindings directly from R. Combined with [treesitter.c](https://github.com/sounkou-bioinfo/treesitter.c), which provides C header parsers, it can be used to rapidly generate declarative bindings.


## How it works

When you call `tcc_compile()`, Rtinycc generates C wrapper functions whose
signature follows the `.Call` convention (`SEXP` in, `SEXP` out). These wrappers
convert R types to C, call the target function, and convert the result back.
TCC compiles them in-memory -- no shared library is written to disk and no
`R_init_*` registration is needed.

After `tcc_relocate()`, wrapper pointers are retrieved via `tcc_get_symbol()`,
which internally calls `RC_libtcc_get_symbol()`. That function converts TCC's
raw `void*` into a `DL_FUNC` wrapped with `R_MakeExternalPtrFn` (tagged
`"native symbol"`). The symbol pointer retains its owning `tcc_state`, so the
relocated code remains alive while the symbol is reachable. On the R side,
[`make_callable()`](R/ffi.R) creates a closure that passes this external pointer
to `.Call` (aliased as `.RtinyccCall` to keep `R CMD check` happy). Non-memory
states instead produce one-shot artifacts with `tcc_output_file()` and cannot
expose callable symbol pointers.

The design follows [CFFI's](https://cffi.readthedocs.io/) API-mode pattern:
instead of computing struct layouts and calling conventions in R (ABI-mode,
like Python's ctypes), the generated C code lets TCC handle `sizeof`,
`offsetof`, and argument passing. Rtinycc never replicates platform-specific
layout rules. The wrappers can also link against external shared libraries
whose symbols TCC resolves at relocation time. For background on how this
compares to a libffi approach, see the
[`RSimpleFFI` README](https://github.com/sounkou-bioinfo/RSimpleFFI#readme).

On macOS the configure script strips `-flat_namespace` from TCC's build to
avoid [BUS ERROR issues](https://genomic.social/@bioinfhotep/115765645745231377). Without it, TCC cannot resolve host symbols (e.g.
`RC_free_finalizer`) through the dynamic linker. Rtinycc works around this
with `RC_libtcc_add_host_symbols()`, which registers package-internal C
functions via `tcc_add_symbol()` before relocation. Any new C function
referenced by generated TCC code must be added there.

On Windows, the `configure.win` script generates a UCRT-backed `msvcrt.def` so
TinyCC resolves CRT symbols against `ucrtbase.dll` (R 4.2+ uses UCRT).


Ownership semantics are explicit. Pointers from `tcc_malloc()` are tagged `rtinycc_owned` and can be released with `tcc_free()` (or by their R finalizer). Generated struct constructors use a struct-specific tag
(`struct_<name>`) and host-allocated storage with `RC_owned_native_finalizer`; free them with `struct_<name>_free()`, not `tcc_free()`. Pointers from `tcc_data_ptr()` are
tagged `rtinycc_borrowed` and are never freed by Rtinycc. Array returns are copied into a fresh R vector; set `free = TRUE` only when the C function returns a `malloc`-owned buffer.

## Installation

``` r
install.packages(
      'Rtinycc', 
        repos = c('https://sounkou-bioinfo.r-universe.dev', 
                  'https://cloud.r-project.org')
        )
```

## Usage

### CLI

The CLI interface compiles C source files to standalone executables using the bundled TinyCC toolchain.

```{r cli}
library(Rtinycc)

src <- system.file("c_examples", "forty_two.c", package = "Rtinycc")
exe <- tempfile()
tcc_run_cli(c(
  "-B", tcc_prefix(),
  paste0("-I", tcc_include_paths()),
  paste0("-L", tcc_lib_paths()),
  src, "-o", exe
))
Sys.chmod(exe, mode = "0755")
system2(exe, stdout = TRUE)
```

For in-memory workflows, prefer libtcc instead.

### In-memory compilation with libtcc

We can compile and call C functions entirely in memory. This is the simplest path for quick JIT compilation.

```{r in-memory}
state <- tcc_state(output = "memory")
tcc_compile_string(state, "int forty_two(){ return 42; }")
tcc_relocate(state)
tcc_call_symbol(state, "forty_two", return = "int")
```

For low-level pointer-style calls, `tcc_call_symbol()` can also use an
R `.C()`-like convention: call a `void` C routine with pointers to copied
argument buffers, then return a list of the modified values.

```{r in-memory-dotc}
state <- tcc_state(output = "memory")
tcc_compile_string(state, "void add_one(int *x) { x[0] += 1; }")
tcc_relocate(state)
tcc_call_symbol(state, "add_one", x = as.integer(41))
```

The lower-level API gives full control over include paths, libraries, and the R C API. Using `#define _Complex` as a workaround for TCC's lack of [complex type support](https://lists.gnu.org/archive/html/tinycc-devel/2022-04/msg00020.html), we can link against R's headers and call into `libR`.

```{r call-R-C-API}
state <- tcc_state(output = "memory")
tcc_add_include_path(state, R.home("include"))
tcc_add_library_path(state, R.home("lib"))

code <- '
#define _Complex
#include <R.h>
#include <Rinternals.h>

double call_r_sqrt(void) {
  SEXP fn   = PROTECT(Rf_findFun(Rf_install("sqrt"), R_BaseEnv));
  SEXP val  = PROTECT(Rf_ScalarReal(16.0));
  SEXP call = PROTECT(Rf_lang2(fn, val));
  SEXP out  = PROTECT(Rf_eval(call, R_GlobalEnv));
  double res = REAL(out)[0];
  UNPROTECT(4);
  return res;
}
'
tcc_compile_string(state, code)
tcc_relocate(state)
tcc_call_symbol(state, "call_r_sqrt", return = "double")
```

### Pointer utilities

Rtinycc ships a set of typed memory access functions similar to what the [ctypesio](https://cran.r-project.org/package=ctypesio) package offers, but designed around our FFI pointer model. Every scalar C type has a corresponding `tcc_read_*` / `tcc_write_*` pair that operates at a byte offset into any external pointer, so you can walk structs, arrays, and output parameters without writing C helpers.

```{r ffi-utils}
ptr <- tcc_cstring("hello")
tcc_read_cstring(ptr)
tcc_read_bytes(ptr, 5)
tcc_ptr_addr(ptr, hex = TRUE)
tcc_ptr_is_null(ptr)
tcc_free(ptr)
```

Typed reads and writes cover the full scalar range (`i8`/`u8`, `i16`/`u16`, `i32`/`u32`, `i64`/`u64`, `f32`/`f64`) plus pointer dereferencing via `tcc_read_ptr` / `tcc_write_ptr`. All operations use a byte offset and `memcpy` internally for alignment safety.

```{r ffi-typed-rw}
buf <- tcc_malloc(32)
tcc_write_i32(buf, 0L, 42L)
tcc_write_f64(buf, 8L, pi)
tcc_read_i32(buf, offset = 0L)
tcc_read_f64(buf, offset = 8L)
tcc_free(buf)
```

Pointer-to-pointer workflows are supported for C APIs that return values through output parameters.

```{r ptr-to-ptr}
ptr_ref <- tcc_malloc(.Machine$sizeof.pointer %||% 8L)
target <- tcc_malloc(8)
tcc_ptr_set(ptr_ref, target)
tcc_data_ptr(ptr_ref)
tcc_ptr_set(ptr_ref, tcc_null_ptr())
tcc_free(target)
tcc_free(ptr_ref)
```

## Declarative FFI

A declarative interface inspired by [Bun's FFI](https://bun.com/docs/runtime/ffi) sits on top of the lower-level API. We define types explicitly and Rtinycc generates the binding code, compiling it in memory with TCC.

### Type system

The FFI exposes a small set of type mappings between R and C. Conversions are explicit and predictable so callers know when data is shared versus copied.

The scalar type names are C-facing, but the R-side carriers are not all
one-to-one with those C widths:

- `i8`, `i16`, `i32`, `u8`, and `u16` are mediated through R integer scalars
- `u32`, `i64`, `u64`, `f32`, and `f64` are mediated through R numeric
  (`double`) coercion and boxing
- `bool` uses R logical
- `cstring` uses an R character scalar

This means `u32` is routed through `double` to preserve the full unsigned
32-bit range, and `i64` / `u64` are only exact up to `2^53` on the R side.

Array arguments pass R vectors to C with zero copy: `raw` maps to `uint8_t*`, `integer_array` to `int32_t*`, `numeric_array` to `double*`. 

Pointer types include `ptr` (opaque external pointer), `sexp` (pass a `SEXP` directly), and callback signatures like `callback:double(double)`.

Variadic functions are supported in two forms: typed prefix tails (`varargs`) and
bounded dynamic tails (`varargs_types` + `varargs_min`/`varargs_max`). Prefix
mode is the cheaper runtime path because dispatch is by tail arity only;
bounded dynamic mode adds per-call scalar type inference to select a compatible
wrapper. For hot loops, prefer fixed arity first, then prefix variadics with a
tight maximum tail size.


Array returns use `returns = list(type = "integer_array", length_arg = 2, free = TRUE)` to copy the result into a new R vector. The `length_arg` is the 1-based index of the C argument that carries the array length. Set `free = TRUE` when the C function returns a `malloc`-owned buffer.

### Simple functions

```{r ffi-simple}
ffi <- tcc_ffi() |>
  tcc_source("
    int add(int a, int b) { return a + b; }
  ") |>
  tcc_bind(add = list(args = list("i32", "i32"), returns = "i32")) |>
  tcc_compile()

ffi$add(5L, 3L)
```


### Variadic calls (e.g. `Rprintf` style)

Rtinycc supports two ways to bind variadic tails. The legacy approach uses
`varargs` as a typed prefix tail, while the bounded dynamic approach uses
`varargs_types` together with `varargs_min` and `varargs_max`. In the bounded
mode, wrappers are generated across the allowed arity and type combinations,
and runtime dispatch selects the matching wrapper from the scalar tail values
provided at call time.

```{r ffi-variadic-rprintf}
ffi_var <- tcc_ffi() |>
  tcc_header("#include <R_ext/Print.h>") |>
  tcc_source('
    #include <stdarg.h>

    int sum_fmt(int n, ...) {
      va_list ap;
      va_start(ap, n);
      int s = 0;
      for (int i = 0; i < n; i++) s += va_arg(ap, int);
      va_end(ap);
      Rprintf("sum_fmt(%d) = %d\\n", n, s);
      return s;
    }
  ') |>
  tcc_bind(
    Rprintf = list(
      args = list("cstring"),
      variadic = TRUE,
      varargs_types = list("i32"),
      varargs_min = 0L,
      varargs_max = 4L,
      returns = "void"
    ),
    sum_fmt = list(
      args = list("i32"),
      variadic = TRUE,
      varargs_types = list("i32"),
      varargs_min = 0L,
      varargs_max = 4L,
      returns = "i32"
    )
  ) |>
  tcc_compile()

ffi_var$Rprintf("Rprintf via bind: %d + %d = %d\n", 2L, 3L, 5L)
ffi_var$sum_fmt(0L)
ffi_var$sum_fmt(2L, 10L, 20L)
ffi_var$sum_fmt(4L, 1L, 2L, 3L, 4L)
```


### Linking external libraries

We can bind directly to symbols in shared libraries. Here we link against `libm`.

```{r ffi-link}
math <- tcc_ffi() |>
  tcc_library("m") |>
  tcc_bind(
    sqrt  = list(args = list("f64"), returns = "f64"),
    sin   = list(args = list("f64"), returns = "f64"),
    floor = list(args = list("f64"), returns = "f64")
  ) |>
  tcc_compile()

math$sqrt(16.0)
math$sin(pi / 2)
math$floor(3.7)
```

### CUDA through NVRTC on a remote GPU

TinyCC compiles the host adapter, not the CUDA kernel. The adapter links to the
CUDA Driver API and NVRTC; NVRTC compiles CUDA C to PTX, and the driver loads
and launches that PTX. The self-contained demonstration below borrows three R
numeric vectors at the wrapper boundary, copies the inputs to device memory,
launches a runtime-compiled kernel, synchronizes, and copies the result back.

The complete program is collapsed to keep this README navigable. The visible
mirai chunk after it reads that exact knitr source, launches an R daemon on the
rig through the SSH tunnel, executes the program there, and returns the result.
Set `RTINYCC_RUN_SSH_DEMOS=true` while rendering to perform the live run.

<details>
<summary>Show the complete Rtinycc, NVRTC, and CUDA program</summary>

```{r cuda-nvrtc-program, eval=FALSE}
library(Rtinycc)

# This opt-in demonstration is run on an NVIDIA Linux/WSL host. Override the
# paths and architecture when CUDA is installed elsewhere.
cuda_path <- Sys.getenv(
  "RTINYCC_CUDA_DRIVER",
  "/usr/lib/wsl/lib/libcuda.so.1"
)
nvrtc_path <- Sys.getenv(
  "RTINYCC_NVRTC",
  file.path(Sys.getenv("CUDA_HOME", "/usr/local/cuda"), "lib64", "libnvrtc.so")
)
stopifnot(file.exists(cuda_path), file.exists(nvrtc_path))

cuda_arch <- Sys.getenv("RTINYCC_CUDA_ARCH")
if (!nzchar(cuda_arch)) {
  nvidia_smi <- Sys.getenv(
    "RTINYCC_NVIDIA_SMI",
    "/usr/lib/wsl/lib/nvidia-smi"
  )
  compute_capability <- trimws(system2(
    nvidia_smi,
    c("--query-gpu=compute_cap", "--format=csv,noheader"),
    stdout = TRUE
  )[[1L]])
  cuda_arch <- paste0("compute_", gsub("[^0-9]", "", compute_capability))
}
stopifnot(grepl("^compute_[0-9]+$", cuda_arch))

code <- '
#include <stdlib.h>
#include <stdio.h>
#include <string.h>

typedef int CUresult;
typedef int CUdevice;
typedef unsigned long long CUdeviceptr;
typedef struct CUctx_st *CUcontext;
typedef struct CUmod_st *CUmodule;
typedef struct CUfunc_st *CUfunction;
typedef struct CUstream_st *CUstream;

typedef int nvrtcResult;
typedef struct _nvrtcProgram *nvrtcProgram;

extern CUresult cuInit(unsigned int flags);
extern CUresult cuDeviceGet(CUdevice *device, int ordinal);
extern CUresult cuCtxCreate_v2(CUcontext *ctx, unsigned int flags,
                               CUdevice device);
extern CUresult cuCtxDestroy_v2(CUcontext ctx);
extern CUresult cuMemAlloc_v2(CUdeviceptr *ptr, size_t bytes);
extern CUresult cuMemFree_v2(CUdeviceptr ptr);
extern CUresult cuMemcpyHtoD_v2(CUdeviceptr dst, const void *src,
                                size_t bytes);
extern CUresult cuMemcpyDtoH_v2(void *dst, CUdeviceptr src, size_t bytes);
extern CUresult cuModuleLoadData(CUmodule *module, const void *image);
extern CUresult cuModuleUnload(CUmodule module);
extern CUresult cuModuleGetFunction(CUfunction *function, CUmodule module,
                                    const char *name);
extern CUresult cuLaunchKernel(CUfunction function,
                               unsigned int grid_x,
                               unsigned int grid_y,
                               unsigned int grid_z,
                               unsigned int block_x,
                               unsigned int block_y,
                               unsigned int block_z,
                               unsigned int shared_mem_bytes,
                               CUstream stream,
                               void **kernel_params,
                               void **extra);
extern CUresult cuCtxSynchronize(void);

extern nvrtcResult nvrtcCreateProgram(nvrtcProgram *program,
                                      const char *source,
                                      const char *name,
                                      int n_headers,
                                      const char * const *headers,
                                      const char * const *include_names);
extern nvrtcResult nvrtcCompileProgram(nvrtcProgram program,
                                       int n_options,
                                       const char * const *options);
extern nvrtcResult nvrtcGetPTXSize(nvrtcProgram program, size_t *size);
extern nvrtcResult nvrtcGetPTX(nvrtcProgram program, char *ptx);
extern nvrtcResult nvrtcGetProgramLogSize(nvrtcProgram program,
                                          size_t *size);
extern nvrtcResult nvrtcGetProgramLog(nvrtcProgram program, char *log);
extern nvrtcResult nvrtcDestroyProgram(nvrtcProgram *program);

static char rt_cuda_error[8192];

static void rt_cuda_set_error(const char *stage, int code) {
  snprintf(rt_cuda_error, sizeof(rt_cuda_error), "%s failed: %d", stage, code);
}

char *rt_cuda_last_error(void) {
  return rt_cuda_error;
}

static void rt_cuda_set_nvrtc_log(nvrtcProgram program, int code) {
  char *log;
  size_t copy_size;
  size_t size = 0;

  snprintf(rt_cuda_error, sizeof(rt_cuda_error),
           "nvrtcCompileProgram failed: %d", code);
  if (nvrtcGetProgramLogSize(program, &size) != 0 || size <= 1)
    return;
  log = malloc(size);
  if (!log)
    return;
  if (nvrtcGetProgramLog(program, log) != 0) {
    free(log);
    rt_cuda_set_error("nvrtcGetProgramLog", code);
    return;
  }
  copy_size = size - 1;
  if (copy_size >= sizeof(rt_cuda_error))
    copy_size = sizeof(rt_cuda_error) - 1;
  memcpy(rt_cuda_error, log, copy_size);
  rt_cuda_error[copy_size] = 0;
  free(log);
}

int rt_cuda_vec_add(double *x, double *y, double *out, int n) {
  static const char *kernel_source =
    "extern \\"C\\" __global__ void vec_add(const double *x, "
    "const double *y, double *out, int n) {"
    "  int i = (int)(blockIdx.x * blockDim.x + threadIdx.x);"
    "  if (i < n) out[i] = x[i] + y[i];"
    "}";
  const char *options[] = {
    "--gpu-architecture=@CUDA_ARCH@",
    "--std=c++11"
  };
  nvrtcProgram program = 0;
  CUcontext context = 0;
  CUmodule module = 0;
  CUfunction function = 0;
  CUdevice device = 0;
  CUdeviceptr device_x = 0;
  CUdeviceptr device_y = 0;
  CUdeviceptr device_out = 0;
  char *ptx = 0;
  size_t ptx_size = 0;
  size_t bytes;
  void *params[4];
  unsigned int blocks;
  int rc;
  int status = -1;

  rt_cuda_error[0] = 0;
  if (n < 0) {
    strcpy(rt_cuda_error, "negative vector length");
    return -1;
  }
  if (n == 0)
    return 0;
  bytes = (size_t)n * sizeof(double);

  rc = nvrtcCreateProgram(&program, kernel_source, "vec_add.cu",
                          0, 0, 0);
  if (rc != 0) {
    rt_cuda_set_error("nvrtcCreateProgram", rc);
    goto cleanup;
  }
  rc = nvrtcCompileProgram(program, 2, options);
  if (rc != 0) {
    rt_cuda_set_nvrtc_log(program, rc);
    goto cleanup;
  }
  rc = nvrtcGetPTXSize(program, &ptx_size);
  if (rc != 0) {
    rt_cuda_set_error("nvrtcGetPTXSize", rc);
    goto cleanup;
  }
  ptx = malloc(ptx_size);
  if (!ptx) {
    strcpy(rt_cuda_error, "PTX allocation failed");
    goto cleanup;
  }
  rc = nvrtcGetPTX(program, ptx);
  if (rc != 0) {
    rt_cuda_set_error("nvrtcGetPTX", rc);
    goto cleanup;
  }

  rc = cuInit(0);
  if (rc != 0) {
    rt_cuda_set_error("cuInit", rc);
    goto cleanup;
  }
  rc = cuDeviceGet(&device, 0);
  if (rc != 0) {
    rt_cuda_set_error("cuDeviceGet", rc);
    goto cleanup;
  }
  rc = cuCtxCreate_v2(&context, 0, device);
  if (rc != 0) {
    rt_cuda_set_error("cuCtxCreate_v2", rc);
    goto cleanup;
  }
  rc = cuModuleLoadData(&module, ptx);
  if (rc != 0) {
    rt_cuda_set_error("cuModuleLoadData", rc);
    goto cleanup;
  }
  rc = cuModuleGetFunction(&function, module, "vec_add");
  if (rc != 0) {
    rt_cuda_set_error("cuModuleGetFunction", rc);
    goto cleanup;
  }
  rc = cuMemAlloc_v2(&device_x, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemAlloc(x)", rc);
    goto cleanup;
  }
  rc = cuMemAlloc_v2(&device_y, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemAlloc(y)", rc);
    goto cleanup;
  }
  rc = cuMemAlloc_v2(&device_out, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemAlloc(out)", rc);
    goto cleanup;
  }
  rc = cuMemcpyHtoD_v2(device_x, x, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemcpyHtoD(x)", rc);
    goto cleanup;
  }
  rc = cuMemcpyHtoD_v2(device_y, y, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemcpyHtoD(y)", rc);
    goto cleanup;
  }

  params[0] = &device_x;
  params[1] = &device_y;
  params[2] = &device_out;
  params[3] = &n;
  blocks = ((unsigned int)n + 255U) / 256U;
  rc = cuLaunchKernel(function, blocks, 1, 1, 256, 1, 1,
                      0, 0, params, 0);
  if (rc != 0) {
    rt_cuda_set_error("cuLaunchKernel", rc);
    goto cleanup;
  }
  rc = cuCtxSynchronize();
  if (rc != 0) {
    rt_cuda_set_error("cuCtxSynchronize", rc);
    goto cleanup;
  }
  rc = cuMemcpyDtoH_v2(out, device_out, bytes);
  if (rc != 0) {
    rt_cuda_set_error("cuMemcpyDtoH", rc);
    goto cleanup;
  }
  status = 0;

cleanup:
  if (device_out)
    cuMemFree_v2(device_out);
  if (device_y)
    cuMemFree_v2(device_y);
  if (device_x)
    cuMemFree_v2(device_x);
  if (module)
    cuModuleUnload(module);
  if (context)
    cuCtxDestroy_v2(context);
  free(ptx);
  if (program)
    nvrtcDestroyProgram(&program);
  return status;
}
'
code <- sub("@CUDA_ARCH@", cuda_arch, code, fixed = TRUE)

cuda <- tcc_ffi() |>
  tcc_source(code) |>
  tcc_library(cuda_path) |>
  tcc_library(nvrtc_path) |>
  tcc_bind(
    rt_cuda_vec_add = list(
      args = list("numeric_array", "numeric_array", "numeric_array", "i32"),
      returns = "i32"
    ),
    rt_cuda_last_error = list(args = list(), returns = "cstring")
  ) |>
  tcc_compile()

n <- as.integer(Sys.getenv("RTINYCC_CUDA_N", "1000000"))
stopifnot(length(n) == 1L, !is.na(n), n > 0L)
set.seed(1)
x <- runif(n)
y <- runif(n)
out <- numeric(n)

timing <- system.time({
  rc <- cuda$rt_cuda_vec_add(x, y, out, n)
})
if (rc != 0L) {
  stop("CUDA vector add failed: ", cuda$rt_cuda_last_error())
}

expected <- x + y
max_error <- max(abs(out - expected))
result <- list(
  host = Sys.info()[["nodename"]],
  rtinycc_version = as.character(packageVersion("Rtinycc")),
  device = trimws(system2(
    Sys.getenv("RTINYCC_NVIDIA_SMI", "/usr/lib/wsl/lib/nvidia-smi"),
    c("--query-gpu=name", "--format=csv,noheader"),
    stdout = TRUE
  )[[1L]]),
  architecture = cuda_arch,
  n = n,
  status = rc,
  elapsed_seconds_including_nvrtc = unname(timing[["elapsed"]]),
  max_error = max_error,
  first_values = head(out)
)
print(result)
stopifnot(max_error == 0)
invisible(result)
```

</details>

`mirai` supports SSH launch and tunnelling directly. The explicit port matters
here because `ssh_config()` uses the port from its URL rather than the `Port`
value of the `rig` SSH alias. `crew` provides higher-level orchestration on
mirai, while a dedicated mirai `"gpu"` profile is sufficient for this rig.

```{r mirai-cuda-ssh, eval=run_ssh_demos}
library(mirai)

cuda_code <- paste(
  knitr::knit_code$get("cuda-nvrtc-program"),
  collapse = "\n"
)
stopifnot(nzchar(cuda_code))

cuda_result <- tryCatch(
  {
    daemons(
      n = 1L,
      url = local_url(tcp = TRUE),
      remote = ssh_config(
        "ssh://sounkoutoure@localhost:2222",
        tunnel = TRUE
      ),
      .compute = "gpu"
    )
    job <- mirai(
      eval(parse(text = code), envir = new.env(parent = globalenv())),
      code = cuda_code,
      .compute = "gpu"
    )
    job[]
  },
  finally = daemons(0, .compute = "gpu")
)

stopifnot(
  is.list(cuda_result),
  identical(cuda_result$status, 0L),
  identical(cuda_result$max_error, 0)
)
data.frame(
  host = cuda_result$host,
  rtinycc = cuda_result$rtinycc_version,
  device = cuda_result$device,
  architecture = cuda_result$architecture,
  n = cuda_result$n,
  elapsed_seconds = cuda_result$elapsed_seconds_including_nvrtc,
  max_error = cuda_result$max_error,
  check.names = FALSE
)
```

The displayed result is returned by the remote R process. Timing includes NVRTC
compilation, CUDA context creation, device allocation, transfers, launch, and
synchronization; this is a correctness demonstration rather than a steady-state
kernel benchmark.

### Compiler options

Use `tcc_options()` to pass raw TinyCC options in the high-level FFI pipeline.
For low-level states, use `tcc_set_options()` directly.

```{r ffi-options}
ffi_opt_off <- tcc_ffi() |>
  tcc_options("-O0") |>
  tcc_source('
    int opt_macro() {
    #ifdef __OPTIMIZE__
      return 1;
    #else
      return 0;
    #endif
    }
  ') |>
  tcc_bind(opt_macro = list(args = list(), returns = "i32")) |>
  tcc_compile()

ffi_opt_on <- tcc_ffi() |>
  tcc_options(c("-Wall", "-O2")) |>
  tcc_source('
    int opt_macro() {
    #ifdef __OPTIMIZE__
      return 1;
    #else
      return 0;
    #endif
    }
  ') |>
  tcc_bind(opt_macro = list(args = list(), returns = "i32")) |>
  tcc_compile()

ffi_opt_off$opt_macro()
ffi_opt_on$opt_macro()
```
### Working with arrays

R vectors are passed to C with zero copy. Mutations in C are visible in R.

```{r ffi-arrays}
ffi <- tcc_ffi() |>
  tcc_source("
    #include <stdlib.h>
    #include <string.h>

    int64_t sum_array(int32_t* arr, int32_t n) {
      int64_t s = 0;
      for (int i = 0; i < n; i++) s += arr[i];
      return s;
    }

    void bump_first(int32_t* arr) { arr[0] += 10; }

    int32_t* dup_array(int32_t* arr, int32_t n) {
      int32_t* out = malloc(sizeof(int32_t) * n);
      memcpy(out, arr, sizeof(int32_t) * n);
      return out;
    }
  ") |>
  tcc_bind(
    sum_array  = list(args = list("integer_array", "i32"), returns = "i64"),
    bump_first = list(args = list("integer_array"), returns = "void"),
    dup_array  = list(
      args = list("integer_array", "i32"),
      returns = list(type = "integer_array", length_arg = 2, free = TRUE)
    )
  ) |>
  tcc_compile()

x <- as.integer(1:100) # to avoid ALTREP
.Internal(inspect(x))
ffi$sum_array(x, length(x))

# Zero-copy: C mutation reflects in R
ffi$bump_first(x)
x[1]

# Array return: copied into a new R vector, C buffer freed
y <- ffi$dup_array(x, length(x))
y[1]

.Internal(inspect(x))
```
## Advanced FFI features

### Structs and unions

Complex C types are supported declaratively. Use `tcc_struct()` to generate allocation and accessor helpers. Free instances when done.

```{r struct-example}
ffi <- tcc_ffi() |>
  tcc_source('
    #include <math.h>
    struct point { double x; double y; };
    double distance(struct point* a, struct point* b) {
      double dx = a->x - b->x, dy = a->y - b->y;
      return sqrt(dx * dx + dy * dy);
    }
  ') |>
  tcc_library("m") |>
  tcc_struct("point", accessors = c(x = "f64", y = "f64")) |>
  tcc_bind(distance = list(args = list("ptr", "ptr"), returns = "f64")) |>
  tcc_compile()

p1 <- ffi$struct_point_new()
ffi$struct_point_set_x(p1, 0.0)
ffi$struct_point_set_y(p1, 0.0)

p2 <- ffi$struct_point_new()
ffi$struct_point_set_x(p2, 3.0)
ffi$struct_point_set_y(p2, 4.0)

ffi$distance(p1, p2)

ffi$struct_point_free(p1)
ffi$struct_point_free(p2)
```

### Enums

Enums are exposed as helper functions that return integer constants.

```{r enum-example}
ffi <- tcc_ffi() |>
  tcc_source("enum color { RED = 0, GREEN = 1, BLUE = 2 };") |>
  tcc_enum("color", constants = c("RED", "GREEN", "BLUE")) |>
  tcc_compile()

ffi$enum_color_RED()
ffi$enum_color_BLUE()
```

### Bitfields

Bitfields are handled by TCC. Accessors read and write them like normal fields.

```{r bitfield-example}
ffi <- tcc_ffi() |>
  tcc_source("
    struct flags {
      unsigned int active : 1;
      unsigned int level  : 4;
    };
  ") |>
  tcc_struct("flags", accessors = c(active = "u8", level = "u8")) |>
  tcc_compile()

s <- ffi$struct_flags_new()
ffi$struct_flags_set_active(s, 1L)
ffi$struct_flags_set_level(s, 9L)
ffi$struct_flags_get_active(s)
ffi$struct_flags_get_level(s)
ffi$struct_flags_free(s)
```

### Global getters and setters

C globals can be exposed with explicit getter/setter helpers.

```{r globals-example}
ffi <- tcc_ffi() |>
  tcc_source("
    int counter = 7;
    double pi_approx = 3.14159;
  ") |>
  tcc_global("counter", "i32") |>
  tcc_global("pi_approx", "f64") |>
  tcc_compile()

ffi$global_counter_get()
ffi$global_pi_approx_get()
ffi$global_counter_set(42L)
ffi$global_counter_get()
```

### Callbacks

R functions can be registered as C function pointers via `tcc_callback()` and passed to compiled code. Specify a `callback:<signature>` argument in `tcc_bind()` so the trampoline is generated automatically. Call `tcc_callback_close()` when you want deterministic invalidation and earlier release of the preserved R function. The legacy `threadsafe` argument is informational; worker dispatch is selected by `callback_async:<signature>`.

```{r callback-example}
cb <- tcc_callback(function(x) x * x, signature = "double (*)(double)")

code <- '
double apply_fn(double (*fn)(void* ctx, double), void* ctx, double x) {
  return fn(ctx, x);
}
'

ffi <- tcc_ffi() |>
  tcc_source(code) |>
  tcc_bind(
    apply_fn = list(
      args = list("callback:double(double)", "ptr", "f64"),
      returns = "f64"
    )
  ) |>
  tcc_compile()

ffi$apply_fn(cb, tcc_callback_ptr(cb), 7.0)
tcc_callback_close(cb)
```

### Callback errors

If a callback throws an R error, the trampoline catches it, emits a warning, and returns a type-appropriate sentinel instead of unwinding through C. In practice this means NA-like numeric or integer values, `NA` logical, `NULL` for `cstring`, or a null external pointer depending on the declared return type.

```{r callback-error}
cb_err <- tcc_callback(
  function(x) stop("boom"),
  signature = "double (*)(double)"
)

ffi_err <- tcc_ffi() |>
  tcc_source('
    double call_cb_err(double (*cb)(void* ctx, double), void* ctx, double x) {
      return cb(ctx, x);
    }
  ') |>
  tcc_bind(
    call_cb_err = list(
      args = list("callback:double(double)", "ptr", "f64"),
      returns = "f64"
    )
  ) |>
  tcc_compile()

warned <- FALSE
res <- withCallingHandlers(
  ffi_err$call_cb_err(cb_err, tcc_callback_ptr(cb_err), 1.0),
  warning = function(w) {
    warned <<- TRUE
    invokeRestart("muffleWarning")
  }
)
list(warned = warned, result = res)
tcc_callback_close(cb_err)
```

### Async callbacks

For callback dispatch from worker threads, use
`callback_async:<signature>` in `tcc_bind()`. The async callback queue
is initialized automatically at package load.

When a bound function has any `callback_async:` argument, the generated
wrapper automatically runs your C function on a new thread while draining
callbacks on the main R thread. Your C code doesn't need to know about
draining at all — just call the callback as normal and the wrapper handles
the rest. Interrupt requests are deferred until the worker has returned and
been joined, so native stack contexts cannot escape while still in use.

**Void return (fire-and-forget):** the callback is enqueued from any
thread and executed on the main R thread automatically — on Windows via
R's message pump, on Linux/macOS via R's event loop `addInputHandler`.

**Non-void return (synchronous):** the worker thread blocks until the
main R thread executes the callback and returns the real result.
Supported return types: integer variants (`int`, `int32_t`, `i8`, `i16`,
`u8`, `u16`, `u32`, `i64`, `u64`), floating-point (`double`, `float`),
`bool`/`logical`, and pointer (`void*`, `T*`). Wide integer values must remain
within R's exact numeric range (absolute value at most `2^53`).

Async scalar and string payloads are copied into queue tasks. Pointer payloads
copy only the address: C must keep the pointee alive until the callback has
executed. Closing a callback cancels queued work for that callback; registry
slot reuse cannot redirect old work to a newly registered callback.

```{r callback-async}
# Fire-and-forget: void callback accumulated from 100 worker threads
hits <- 0L
cb_async <- tcc_callback(
  function(x) { hits <<- hits + x; NULL },
  signature = "void (*)(int)"
)

code_async <- '
struct task { void (*cb)(void* ctx, int); void* ctx; int value; };

#ifdef _WIN32
#include <windows.h>

static DWORD WINAPI worker(LPVOID data) {
  struct task* t = (struct task*) data;
  t->cb(t->ctx, t->value);
  return 0;
}

int spawn_async(void (*cb)(void* ctx, int), void* ctx, int value) {
  if (!cb || !ctx) return -1;
  struct task t;
  t.cb = cb;
  t.ctx = ctx;
  t.value = value;
  HANDLE th = CreateThread(NULL, 0, worker, &t, 0, NULL);
  if (!th) return -2;
  WaitForSingleObject(th, INFINITE);
  CloseHandle(th);
  return 0;
}
#else
#include <pthread.h>

static void* worker(void* data) {
  struct task* t = (struct task*) data;
  t->cb(t->ctx, t->value);
  return NULL;
}

int spawn_async(void (*cb)(void* ctx, int), void* ctx, int value) {
  if (!cb || !ctx) return -1;
  const int n = 100;
  struct task tasks[100];
  pthread_t th[100];
  for (int i = 0; i < n; i++) {
    tasks[i].cb = cb;
    tasks[i].ctx = ctx;
    tasks[i].value = value;
    if (pthread_create(&th[i], NULL, worker, &tasks[i]) != 0) {
      for (int j = 0; j < i; j++) pthread_join(th[j], NULL);
      return -2;
    }
  }
  for (int i = 0; i < n; i++) pthread_join(th[i], NULL);
  return 0;
}
#endif
'

ffi_async <- tcc_ffi() |>
  tcc_source(code_async)
if (.Platform$OS.type != "windows") {
  ffi_async <- tcc_library(ffi_async, "pthread")
}
ffi_async <- ffi_async |>
  tcc_bind(
    spawn_async = list(
      args = list("callback_async:void(int)", "ptr", "i32"),
      returns = "i32"
    )
  ) |>
  tcc_compile()

rc <- ffi_async$spawn_async(cb_async, tcc_callback_ptr(cb_async), 2L)
hits
tcc_callback_close(cb_async)
```

Non-void return works the same way — the generated wrapper handles the
drain loop transparently:

```{r callback-async-nonvoid}
cb_triple <- tcc_callback(
  function(x) x * 3L,
  signature = "int (*)(int)"
)

# Pure C: the worker calls the sync callback and returns its result.
# No drain logic needed — the generated wrapper handles it.
code_sync <- '
#ifdef _WIN32
#include <windows.h>

struct itask { int (*cb)(void*,int); void* ctx; int in; int out; };

static DWORD WINAPI iworker(LPVOID p) {
  struct itask* t = (struct itask*)p;
  t->out = t->cb(t->ctx, t->in);
  return 0;
}
int run_worker(int (*cb)(void*,int), void* ctx, int x) {
  struct itask t;
  t.cb = cb; t.ctx = ctx; t.in = x; t.out = -1;
  HANDLE th = CreateThread(NULL, 0, iworker, &t, 0, NULL);
  if (!th) return -1;
  WaitForSingleObject(th, INFINITE);
  CloseHandle(th);
  return t.out;
}
#else
#include <pthread.h>

struct itask { int (*cb)(void*,int); void* ctx; int in; int out; };

static void* iworker(void* p) {
  struct itask* t = (struct itask*)p;
  t->out = t->cb(t->ctx, t->in);
  return NULL;
}
int run_worker(int (*cb)(void*,int), void* ctx, int x) {
  struct itask t;
  t.cb = cb; t.ctx = ctx; t.in = x; t.out = -1;
  pthread_t th;
  if (pthread_create(&th, NULL, iworker, &t) != 0) return -1;
  pthread_join(th, NULL);
  return t.out;
}
#endif
'

ffi_sync <- tcc_ffi() |>
  tcc_source(code_sync)
if (.Platform$OS.type != "windows") {
  ffi_sync <- tcc_library(ffi_sync, "pthread")
}
ffi_sync <- ffi_sync |>
  tcc_bind(
    run_worker = list(args = list("callback_async:int(int)", "ptr", "i32"), returns = "i32")
  ) |>
  tcc_compile()

ffi_sync$run_worker(cb_triple, tcc_callback_ptr(cb_triple), 7L)  # 21
tcc_callback_close(cb_triple)
```

### SQLite: a complete example

This example ties together external library linking, callbacks, and pointer dereferencing. We open an in-memory SQLite database, execute queries, and collect rows through an R callback that reads `char**` arrays using `tcc_read_ptr` and `tcc_read_cstring`.

```{r ffi-sqlite, eval=TRUE}
ptr_size <- .Machine$sizeof.pointer

read_string_array <- function(ptr, n) {
  vapply(seq_len(n), function(i) {
    tcc_read_cstring(tcc_read_ptr(ptr, (i - 1L) * ptr_size))
  }, "")
}

cb <- tcc_callback(
  function(argc, argv, cols) {
    values <- read_string_array(argv, argc)
    names  <- read_string_array(cols, argc)
    cat(paste(names, values, sep = " = ", collapse = ", "), "\n")
    0L
  },
  signature = "int (*)(int, char **, char **)"
)

sqlite <- tcc_ffi() |>
  tcc_header("#include <sqlite3.h>") |>
  tcc_library("sqlite3") |>
  tcc_source('
    void* open_db() {
      sqlite3* db = NULL;
      sqlite3_open(":memory:", &db);
      return db;
    }
    int close_db(void* db) {
      return sqlite3_close((sqlite3*)db);
    }
  ') |>
  tcc_bind(
    open_db  = list(args = list(), returns = "ptr"),
    close_db = list(args = list("ptr"), returns = "i32"),
    sqlite3_libversion = list(args = list(), returns = "cstring"),
    sqlite3_exec = list(
      args = list("ptr", "cstring", "callback:int(int, char **, char **)", "ptr", "ptr"),
      returns = "i32"
    )
  ) |>
  tcc_compile()

sqlite$sqlite3_libversion()

db <- sqlite$open_db()
sqlite$sqlite3_exec(db, "CREATE TABLE t (id INTEGER, name TEXT);", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$sqlite3_exec(db, "INSERT INTO t VALUES (1, 'hello'), (2, 'world');", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$sqlite3_exec(db, "SELECT * FROM t;", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$close_db(db)
tcc_callback_close(cb)
```

### Stackless C Protothreads: streaming BCF/VCF records with htslib

Rtinycc bundles Adam Dunkels' [Protothreads](http://dunkels.com/adam/pt/) library, allowing you to JIT-compile cross-platform native iterators that yield back to R. This demo binds [htslib](https://www.htslib.org/) using a protothread:
R resumes the C reader until the next `bcf1_t` is ready, then copies the current fields into a regular R list. The C logic never blocks R's event loop, and R objects are created only after control returns safely to the main R stack.

**Important Protothreads Limitations:**
Because protothreads are stackless (implemented using Duff's device macro expansions), you must observe a few strict rules in your C code:

1. **Local variables are NOT preserved** across `PT_YIELD` calls. Store state in your context struct instead.
2. **No nested blocking**: You cannot yield from inside a nested C function call.
3. **No switch statements**: Because Protothreads uses `switch` internally, you cannot use `switch` within a `PT_THREAD` block.

The demo uses plain VCF text, opened directly by htslib through the same API as BCF.
```{r ffi-bcf-coroutine-run, eval=nzchar(Sys.which("pkg-config")) && system2("pkg-config", c("--exists", "htslib")) == 0L}
source("scripts/demo-streaming-bcf-reader-ffi.R")
run_streaming_bcf_demo()
```

The full demo source is foldable below.

<details>
<summary>Click to show the complete <code>scripts/demo-streaming-bcf-reader-ffi.R</code> script</summary>

```{r ffi-bcf-coroutine-demo-source, echo=FALSE, results='asis'}
bcf_demo_path <- "scripts/demo-streaming-bcf-reader-ffi.R"
bcf_demo_source <- paste(readLines(bcf_demo_path, warn = FALSE), collapse = "\n")
knitr::asis_output(knitr::engine_output(
  list(
    echo = TRUE,
    eval = FALSE,
    engine = "r",
    class.source = "rtinycc-demo",
    prompt = FALSE,
    comment = "",
    tidy = FALSE,
    highlight = TRUE
  ),
  bcf_demo_source,
  ""
))
```

</details>

The embedded htslib Protothreads C source is extracted from the script and rendered
with the same Rtinycc C-code display helper used by the vignettes.

<details>
<summary>Click to show only the embedded C Protothreads / htslib source</summary>

```{r ffi-bcf-coroutine-c-source, echo=FALSE, results='asis'}
bcf_demo_lines <- readLines(bcf_demo_path, warn = FALSE)
bcf_c_start <- grep("^  code <- '$", bcf_demo_lines)
bcf_c_end <- NA_integer_
if (length(bcf_c_start) == 1L) {
  bcf_c_end <- which(seq_along(bcf_demo_lines) > bcf_c_start & bcf_demo_lines == "'")[1]
}
if (length(bcf_c_start) == 1L && length(bcf_c_end) == 1L && !is.na(bcf_c_end)) {
  bcf_c_source <- paste(bcf_demo_lines[(bcf_c_start + 1L):(bcf_c_end - 1L)], collapse = "\n")
  Rtinycc:::rtinycc_c_block(bcf_c_source)
} else {
  knitr::asis_output("_Could not extract embedded C source from the demo script._")
}
```

</details>

This points toward a useful separate package idea: a small, generic
Rtinycc-based C coroutine/Protothreads layer for R.  Such a package could provide reusable
pieces like `coro_new()`, `coro_resume()`, finalizer-backed native handles,
status/error conventions, and helpers for writing streaming bindings around C
libraries such as htslib, SQLite, parsers, decompression libraries, or network
clients.  R would see ordinary iterators; C would keep natural blocking or
recursive control flow without rewriting everything as a heap-allocated state
machine.

## Header parsing with treesitter.c

For header-driven bindings, we use `treesitter.c` to parse function signatures and generate binding specifications automatically. For struct, enum, and global helpers, `tcc_generate_bindings()` handles the code generation.

The default mapper is conservative for pointers: `char*` is treated as `ptr` because C does not guarantee NUL-terminated strings. If you know a parameter is a C string, provide a custom mapper that returns `cstring` for that type.

```{r ffi-treesitter}
header <- '
double sqrt(double x);
double sin(double x);
struct point { double x; double y; };
enum status { OK = 0, ERROR = 1 };
int global_counter;
'

tcc_treesitter_functions(header)
tcc_treesitter_structs(header)
tcc_treesitter_enums(header)
tcc_treesitter_globals(header)

# Bind parsed functions to libm
symbols <- tcc_treesitter_bindings(header)
math <- tcc_link("m", symbols = symbols)
math$sqrt(16.0)

# Generate struct/enum/global helpers
ffi <- tcc_ffi() |>
  tcc_source(header) |>
  tcc_generate_bindings(
    header,
    functions = FALSE, structs = TRUE,
    enums = TRUE, globals = TRUE
  ) |>
  tcc_compile()

ffi$struct_point_new()
ffi$enum_status_OK()
ffi$global_global_counter_get()
```

## io_uring Demo

`CSV` parser using [`io_uring`](https://en.wikipedia.org/wiki/Io_uring) on linux

```{r io_uring-demo, eval=TRUE}
if (Sys.info()[["sysname"]] == "Linux") {
  c_file <- system.file("c_examples", "io_uring_csv.c", package = "Rtinycc")

  n_rows <- 20000L
  n_cols <- 8L
  block_size <- 1024L * 1024L

  set.seed(42)
  tmp_csv <- tempfile("rtinycc_io_uring_readme_", fileext = ".csv")
  on.exit(unlink(tmp_csv), add = TRUE)

  mat <- matrix(runif(n_rows * n_cols), ncol = n_cols)
  df <- as.data.frame(mat)
  names(df) <- paste0("V", seq_len(n_cols))
  utils::write.table(df, file = tmp_csv, sep = ",", row.names = FALSE, col.names = TRUE, quote = FALSE)
  csv_size_mb <- as.double(file.info(tmp_csv)$size) / 1024^2
  message(sprintf("CSV size: %.2f MB", csv_size_mb))

  io_uring_src <- paste(readLines(c_file, warn = FALSE), collapse = "\n")

  ffi <- tcc_ffi() |>
    tcc_source(io_uring_src) |>
    tcc_bind(
      csv_table_read = list(
        args = list("cstring", "i32", "i32"),
        returns = "sexp"
      ),
      csv_table_io_uring = list(
        args = list("cstring", "i32", "i32"),
        returns = "sexp"
      )
    ) |>
    tcc_compile()

  baseline <- utils::read.table(tmp_csv, sep = ",", header = TRUE)
  c_tbl <- ffi$csv_table_read(tmp_csv, block_size, n_cols)
  uring_tbl <- ffi$csv_table_io_uring(tmp_csv, block_size, n_cols)
  vroom_tbl <- vroom::vroom(
    tmp_csv,
    delim = ",",
    altrep = FALSE,
    col_types = vroom::cols(.default = "d"),
    progress = FALSE,
    show_col_types = FALSE
  )

  stopifnot(
    identical(dim(c_tbl), dim(baseline)),
    identical(dim(uring_tbl), dim(baseline)),
    identical(dim(vroom_tbl), dim(baseline)),
    isTRUE(all.equal(c_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE)),
    isTRUE(all.equal(uring_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE)),
    isTRUE(all.equal(vroom_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE))
  )

  timings <- bench::mark(
    read_table_df = {
      x <- utils::read.table(tmp_csv, sep = ",", header = TRUE)
      nrow(x)
    },
    vroom_df_altrep_false = {
      x <- vroom::vroom(
        tmp_csv,
        delim = ",",
        altrep = FALSE,
        col_types = vroom::cols(.default = "d"),
        progress = FALSE,
        show_col_types = FALSE
      )
      nrow(x)
    },
    vroom_df_altrep_false_mat = {
      x <- vroom::vroom(
        tmp_csv,
        delim = ",",
        altrep = FALSE,
        col_types = vroom::cols(.default = "d"),
        progress = FALSE,
        show_col_types = FALSE
      )
      x <- as.matrix(x)
      nrow(x)
    },
    c_read_df = {
      x <- ffi$csv_table_read(tmp_csv, block_size, n_cols)
      nrow(x)
    },
    io_uring_df = {
      x <- ffi$csv_table_io_uring(tmp_csv, block_size, n_cols)
      nrow(x)
    },
    iterations = 2,
    memory = TRUE
  )

  
  print(timings)
  
  plot(timings, type = "boxplot") + bench::scale_x_bench_time(base = NULL)
}
```

## Known limitations

### `_Complex` types

TCC does not support C99 `_Complex` types. Generated code works around this with `#define _Complex`, which suppresses the keyword. Apply the same workaround in your own `tcc_source()` code when headers pull in complex types.

### 64-bit integer precision

R represents `i64` and `u64` values as `double`, which loses precision beyond $2^{53}$. Values that differ only past that threshold become indistinguishable.

```{r limits-int64}
sprintf("2^53:     %.0f", 2^53)
sprintf("2^53 + 1: %.0f", 2^53 + 1)
identical(2^53, 2^53 + 1)
```

For exact 64-bit arithmetic, keep values in C-allocated storage and manipulate them through pointers.

### Nested structs

Named nested struct fields can now be declared explicitly with `struct:<name>`.
Getters return borrowed nested views and setters copy bytes from another struct
object of the declared nested type.

```{r limits-nested}
ffi <- tcc_ffi() |>
  tcc_source('
    struct inner { int a; };
    struct outer { struct inner in; };
  ') |>
  tcc_struct("inner", accessors = c(a = "i32")) |>
  tcc_struct("outer", accessors = list(`in` = "struct:inner")) |>
  tcc_compile()

outer <- ffi$struct_outer_new()
inner <- ffi$struct_inner_new()
inner <- ffi$struct_inner_set_a(inner, 42L)
outer <- ffi$struct_outer_set_in(outer, inner)
inner_view <- ffi$struct_outer_get_in(outer)
ffi$struct_inner_get_a(inner_view)
ffi$struct_inner_free(inner)
ffi$struct_outer_free(outer)
```

For treesitter-generated bindings, nested struct fields inside structs still
fall back to ptr-like accessors. If you want a borrowed nested view plus a
copy-in setter, declare the nested field explicitly with `struct:<name>`.

Bitfields are separate from ordinary addressable fields. They use scalar helper
accessors, but `field_addr()` and `container_of()` reject bitfield members.

```{r bitfield-nested-note, error=TRUE}
ffi <- tcc_ffi() |>
  tcc_source('struct flags { unsigned int flag : 1; };') |>
  tcc_struct(
    "flags",
    accessors = list(flag = list(type = "u8", bitfield = TRUE, width = 1))
  )

tcc_field_addr(ffi, "flags", "flag")
tcc_container_of(ffi, "flags", "flag")
```

### Array fields in structs

Array fields require the `list(type = ..., size = N, array = TRUE)` syntax in `tcc_struct()`, which generates element-wise accessors.

```{r limits-arrays}
ffi <- tcc_ffi() |>
  tcc_source('struct buf { unsigned char data[16]; };') |>
  tcc_struct("buf", accessors = list(
    data = list(type = "u8", size = 16, array = TRUE)
  )) |>
  tcc_compile()

b <- ffi$struct_buf_new()
ffi$struct_buf_set_data_elt(b, 0L, 0xCAL)
ffi$struct_buf_set_data_elt(b, 1L, 0xFEL)
ffi$struct_buf_get_data_elt(b, 0L)
ffi$struct_buf_get_data_elt(b, 1L)
ffi$struct_buf_free(b)
```

## Serialization and fork safety

Compiled FFI objects are fork-safe: `parallel::mclapply()` and other `fork()`-based parallelism work out of the box because TCC's compiled code lives in memory mappings that survive `fork()` via copy-on-write.

Serialization is also supported. Each `tcc_compiled` object stores its FFI recipe internally, so after `saveRDS()` / `readRDS()` (or `serialize()` / `unserialize()`), the first `$` access detects the dead TCC state pointer and recompiles transparently.

```{r serialize-example}
ffi <- tcc_ffi() |>
  tcc_source("int square(int x) { return x * x; }") |>
  tcc_bind(square = list(args = list("i32"), returns = "i32")) |>
  tcc_compile()

ffi$square(7L)

tmp <- tempfile(fileext = ".rds")
saveRDS(ffi, tmp)
ffi2 <- readRDS(tmp)
unlink(tmp)

# Auto-recompiles on first access
ffi2$square(7L)
```

For explicit control, use `tcc_recompile()`. Note that raw `tcc_state` objects and bare pointers from `tcc_malloc()` do not carry a recipe and remain dead after deserialization.

## Performance and benchmarking

`Rtinycc` is optimized for fast in-process compilation and convenient FFI
workflows, not for winning every microbenchmark against a conventional
precompiled `.Call()` shared library.

In practice, the usual pattern is:

- `Rtinycc` compiles tiny modules very quickly
- a regular `.Call()` module can have lower minimal per-call overhead
- array-oriented inputs can be zero-copy for already-materialized R vectors,
  but ALTREP inputs may materialize when C pointer access is requested
- return paths that copy native buffers back into fresh R vectors make that
  copy cost visible

If you want the benchmark details rather than the high-level summary, see the
`Compilation and Call Overhead` vignette.

## License

GPL-3

## References

- [TinyCC](https://github.com/TinyCC/tinycc)
- [Bun's FFI](https://bun.com/docs/runtime/ffi)
- [CFFI](https://cffi.readthedocs.io/)
- [RSimpleFFI](https://github.com/sounkou-bioinfo/RSimpleFFI#readme)
- [CSlug](https://cslug.readthedocs.io/en/latest/)
