---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
has_callme <- requireNamespace("callme", quietly = TRUE)
has_bench <- requireNamespace("bench", quietly = TRUE)
```

# Rtinycc

 Builds `TinyCC` `Cli` and Library For `C` Scripting in `R`

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rtinycc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rtinycc/actions/workflows/R-CMD-check.yaml)[![Rtinycc status badge](https://sounkou-bioinfo.r-universe.dev/Rtinycc/badges/version)](https://sounkou-bioinfo.r-universe.dev/Rtinycc)
<!-- badges: end -->

## Abstract

Rtinycc is an R interface to [TinyCC](https://github.com/TinyCC/tinycc), providing both CLI access and a libtcc-backed in-memory compiler. It includes an FFI inspired by [Bun's FFI](https://bun.com/docs/runtime/ffi) for binding C symbols with predictable type conversions and pointer utilities. The package works on unix-alikes and Windows and focuses on embedding TinyCC and enabling JIT-compiled bindings directly from R. Combined with [treesitter.c](https://github.com/sounkou-bioinfo/treesitter.c), which provides C header parsers, it can be used to rapidly generate declarative bindings.


## How it works

When you call `tcc_compile()`, Rtinycc generates C wrapper functions whose
signature follows the `.Call` convention (`SEXP` in, `SEXP` out). These wrappers
convert R types to C, call the target function, and convert the result back.
TCC compiles them in-memory -- no shared library is written to disk and no
`R_init_*` registration is needed.

After `tcc_relocate()`, wrapper pointers are retrieved via `tcc_get_symbol()`,
which internally calls `RC_libtcc_get_symbol()`. That function converts TCC's
raw `void*` into a `DL_FUNC` wrapped with `R_MakeExternalPtrFn` (tagged
`"native symbol"`). On the R side, [`make_callable()`](R/ffi.R) creates a
closure that passes this external pointer to `.Call` (aliased as `.RtinyccCall`
to keep `R CMD check` happy).

The design follows [CFFI's](https://cffi.readthedocs.io/) API-mode pattern:
instead of computing struct layouts and calling conventions in R (ABI-mode,
like Python's ctypes), the generated C code lets TCC handle `sizeof`,
`offsetof`, and argument passing. Rtinycc never replicates platform-specific
layout rules. The wrappers can also link against external shared libraries
whose symbols TCC resolves at relocation time. For background on how this
compares to a libffi approach, see the
[`RSimpleFFI` README](https://github.com/sounkou-bioinfo/RSimpleFFI#readme).

On macOS the configure script strips `-flat_namespace` from TCC's build to
avoid [BUS ERROR issues](https://genomic.social/@bioinfhotep/115765645745231377). Without it, TCC cannot resolve host symbols (e.g.
`RC_free_finalizer`) through the dynamic linker. Rtinycc works around this
with `RC_libtcc_add_host_symbols()`, which registers package-internal C
functions via `tcc_add_symbol()` before relocation. Any new C function
referenced by generated TCC code must be added there.

On Windows, the `configure.win` script generates a UCRT-backed `msvcrt.def` so
TinyCC resolves CRT symbols against `ucrtbase.dll` (R 4.2+ uses UCRT).


Ownership semantics are explicit. Pointers from `tcc_malloc()` are tagged `rtinycc_owned` and can be released with `tcc_free()` (or by their R finalizer). Generated struct constructors use a struct-specific tag
(`struct_<name>`) with an `RC_free_finalizer`; free them with `struct_<name>_free()`, not `tcc_free()`. Pointers from `tcc_data_ptr()` are
tagged `rtinycc_borrowed` and are never freed by Rtinycc. Array returns are copied into a fresh R vector; set `free = TRUE` only when the C function returns a `malloc`-owned buffer.

## Installation

``` r
install.packages(
      'Rtinycc', 
        repos = c('https://sounkou-bioinfo.r-universe.dev', 
                  'https://cloud.r-project.org')
        )
```

## Usage

### CLI

The CLI interface compiles C source files to standalone executables using the bundled TinyCC toolchain.

```{r cli}
library(Rtinycc)

src <- system.file("c_examples", "forty_two.c", package = "Rtinycc")
exe <- tempfile()
tcc_run_cli(c(
  "-B", tcc_prefix(),
  paste0("-I", tcc_include_paths()),
  paste0("-L", tcc_lib_paths()),
  src, "-o", exe
))
Sys.chmod(exe, mode = "0755")
system2(exe, stdout = TRUE)
```

For in-memory workflows, prefer libtcc instead.

### In-memory compilation with libtcc

We can compile and call C functions entirely in memory. This is the simplest path for quick JIT compilation.

```{r in-memory}
state <- tcc_state(output = "memory")
tcc_compile_string(state, "int forty_two(){ return 42; }")
tcc_relocate(state)
tcc_call_symbol(state, "forty_two", return = "int")
```

The lower-level API gives full control over include paths, libraries, and the R C API. Using `#define _Complex` as a workaround for TCC's lack of [complex type support](https://mail.gnu.org/archive/html/tinycc-devel/2022-04/msg00020.html), we can link against R's headers and call into `libR`.

```{r call-R-C-API}
state <- tcc_state(output = "memory")
tcc_add_include_path(state, R.home("include"))
tcc_add_library_path(state, R.home("lib"))

code <- '
#define _Complex
#include <R.h>
#include <Rinternals.h>

double call_r_sqrt(void) {
  SEXP fn   = PROTECT(Rf_findFun(Rf_install("sqrt"), R_BaseEnv));
  SEXP val  = PROTECT(Rf_ScalarReal(16.0));
  SEXP call = PROTECT(Rf_lang2(fn, val));
  SEXP out  = PROTECT(Rf_eval(call, R_GlobalEnv));
  double res = REAL(out)[0];
  UNPROTECT(4);
  return res;
}
'
tcc_compile_string(state, code)
tcc_relocate(state)
tcc_call_symbol(state, "call_r_sqrt", return = "double")
```

### Pointer utilities

Rtinycc ships a set of typed memory access functions similar to what the [ctypesio](https://cran.r-project.org/package=ctypesio) package offers, but designed around our FFI pointer model. Every scalar C type has a corresponding `tcc_read_*` / `tcc_write_*` pair that operates at a byte offset into any external pointer, so you can walk structs, arrays, and output parameters without writing C helpers.

```{r ffi-utils}
ptr <- tcc_cstring("hello")
tcc_read_cstring(ptr)
tcc_read_bytes(ptr, 5)
tcc_ptr_addr(ptr, hex = TRUE)
tcc_ptr_is_null(ptr)
tcc_free(ptr)
```

Typed reads and writes cover the full scalar range (`i8`/`u8`, `i16`/`u16`, `i32`/`u32`, `i64`/`u64`, `f32`/`f64`) plus pointer dereferencing via `tcc_read_ptr` / `tcc_write_ptr`. All operations use a byte offset and `memcpy` internally for alignment safety.

```{r ffi-typed-rw}
buf <- tcc_malloc(32)
tcc_write_i32(buf, 0L, 42L)
tcc_write_f64(buf, 8L, pi)
tcc_read_i32(buf, offset = 0L)
tcc_read_f64(buf, offset = 8L)
tcc_free(buf)
```

Pointer-to-pointer workflows are supported for C APIs that return values through output parameters.

```{r ptr-to-ptr}
ptr_ref <- tcc_malloc(.Machine$sizeof.pointer %||% 8L)
target <- tcc_malloc(8)
tcc_ptr_set(ptr_ref, target)
tcc_data_ptr(ptr_ref)
tcc_ptr_set(ptr_ref, tcc_null_ptr())
tcc_free(target)
tcc_free(ptr_ref)
```

## Declarative FFI

A declarative interface inspired by [Bun's FFI](https://bun.com/docs/runtime/ffi) sits on top of the lower-level API. We define types explicitly and Rtinycc generates the binding code, compiling it in memory with TCC.

### Type system

The FFI exposes a small set of type mappings between R and C. Conversions are explicit and predictable so callers know when data is shared versus copied.

Scalar types map one-to-one: `i8`, `i16`, `i32`, `i64` (integers); `u8`, `u16`, `u32`, `u64` (unsigned); `f32`, `f64` (floats); `bool` (logical); `cstring` (NUL-terminated string). 

Array arguments pass R vectors to C with zero copy: `raw` maps to `uint8_t*`, `integer_array` to `int32_t*`, `numeric_array` to `double*`. 

Pointer types include `ptr` (opaque external pointer), `sexp` (pass a `SEXP` directly), and callback signatures like `callback:double(double)`.

Variadic functions are supported in two forms: typed prefix tails (`varargs`) and
bounded dynamic tails (`varargs_types` + `varargs_min`/`varargs_max`). Prefix
mode is the cheaper runtime path because dispatch is by tail arity only;
bounded dynamic mode adds per-call scalar type inference to select a compatible
wrapper. For hot loops, prefer fixed arity first, then prefix variadics with a
tight maximum tail size.


Array returns use `returns = list(type = "integer_array", length_arg = 2, free = TRUE)` to copy the result into a new R vector. The `length_arg` is the 1-based index of the C argument that carries the array length. Set `free = TRUE` when the C function returns a `malloc`-owned buffer.

### Simple functions

```{r ffi-simple}
ffi <- tcc_ffi() |>
  tcc_source("
    int add(int a, int b) { return a + b; }
  ") |>
  tcc_bind(add = list(args = list("i32", "i32"), returns = "i32")) |>
  tcc_compile()

ffi$add(5L, 3L)
```

### Compilation and call overhead

To get a rough sense of cost, it is useful to compare `Rtinycc` against a
conventional `.Call()` shared-library workflow. The snippet below uses the
[`callme`](https://cran.r-project.org/package=callme) package as the baseline.
That means `callme` is compiling through `R CMD SHLIB` with the platform
compiler toolchain (`gcc`/`clang` on the usual Unix-like targets), so better
steady-state optimization than TinyCC is expected.

Three small cases are enough to make the main point:

- `noop()`: takes nothing and returns nothing
- `fill_rand(out, n)`: fills a caller-provided numeric vector in place
- `rand_unif(n)`: generates `n` random doubles

The `fill_rand()` case is the fairer array-oriented comparison:

- `Rtinycc` uses the `numeric_array` input path and borrows the backing R buffer
- `callme` writes into `REAL(vec)` directly

The `rand_unif()` case then isolates the extra copy path:

- `callme` allocates the final R vector directly with the R C API
- `Rtinycc` returns a native `double*`, and the generated wrapper copies that
  buffer into an R vector

```{r ffi-overhead, eval = has_callme && has_bench}
rtinycc_code <- '
  #include <R.h>
  #include <Rinternals.h>
  #include <Rmath.h>
  #include <stdlib.h>

  void noop(void) {}

  void fill_rand(double* out, int n) {
    if (n < 0) {
      Rf_error("n must be non-negative");
    }

    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      out[i] = unif_rand();
    }
    PutRNGstate();
  }

  double* rand_unif(int n) {
    if (n < 0) {
      Rf_error("n must be non-negative");
    }
    if (n == 0) {
      return (double*) malloc(sizeof(double));
    }

    double *out = (double*) malloc(sizeof(double) * (size_t) n);
    if (!out) {
      Rf_error("malloc failed");
    }

    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      out[i] = unif_rand();
    }
    PutRNGstate();
    return out;
  }
'

callme_code <- '
  #include <R.h>
  #include <Rinternals.h>
  #include <Rmath.h>

  SEXP noop(void) {
    return R_NilValue;
  }

  SEXP fill_rand(SEXP out_, SEXP n_) {
    int n = asInteger(n_);
    if (n < 0) {
      Rf_error("n must be non-negative");
    }

    if (TYPEOF(out_) != REALSXP) {
      Rf_error("out must be a numeric vector");
    }

    if (XLENGTH(out_) < n) {
      Rf_error("out is shorter than n");
    }

    double *out = REAL(out_);
    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      out[i] = unif_rand();
    }
    PutRNGstate();

    return out_;
  }

  SEXP rand_unif(SEXP n_) {
    int n = asInteger(n_);
    if (n < 0) {
      Rf_error("n must be non-negative");
    }

    SEXP out = PROTECT(allocVector(REALSXP, n));
    double *ptr = REAL(out);

    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      ptr[i] = unif_rand();
    }
    PutRNGstate();

    UNPROTECT(1);
    return out;
  }
'

build_rtinycc_module <- function() {
  tcc_ffi() |>
    tcc_source(rtinycc_code) |>
    tcc_bind(
      noop = list(args = list(), returns = "void"),
      fill_rand = list(args = list("numeric_array", "i32"), returns = "void"),
      rand_unif = list(
        args = list("i32"),
        returns = list(type = "numeric_array", length_arg = 1, free = TRUE)
      )
    ) |>
    tcc_compile()
}

build_callme_module <- function() {
  before <- names(getLoadedDLLs())
  mod <- callme::compile(callme_code, env = NULL, verbosity = 0)
  dlls <- getLoadedDLLs()
  new_names <- setdiff(names(dlls), before)
  new_names <- new_names[startsWith(new_names, "callme_")]
  attr(mod, "dll_paths") <- unname(vapply(
    dlls[new_names],
    function(x) x[["path"]],
    character(1)
  ))
  mod
}

unload_callme_dlls <- function(dll_paths) {
  dll_paths <- rev(unique(dll_paths))
  if (is.null(dll_paths) || !length(dll_paths)) {
    return(invisible(NULL))
  }
  for (dll_path in dll_paths) {
    if (is.character(dll_path) && nzchar(dll_path) && file.exists(dll_path)) {
      try(dyn.unload(dll_path), silent = TRUE)
    }
  }
  invisible(NULL)
}

build_and_dispose_callme_module <- function() {
  mod <- build_callme_module()
  dll_paths <- attr(mod, "dll_paths", exact = TRUE)
  rm(mod)
  gc()
  unload_callme_dlls(dll_paths)
  invisible(NULL)
}

with_benchmark_modules <- function(fun) {
  rt_mod <- build_rtinycc_module()
  cm_mod <- build_callme_module()
  dll_paths <- attr(cm_mod, "dll_paths", exact = TRUE)

  on.exit({
    rm(rt_mod, cm_mod)
    gc()
    unload_callme_dlls(dll_paths)
  }, add = TRUE)

  fun(rt_mod, cm_mod)
}

median_elapsed <- function(expr, times = 3L) {
  expr <- substitute(expr)
  env <- parent.frame()
  stats::median(replicate(
    times,
    {
      gc()
      t0 <- proc.time()[["elapsed"]]
      eval(expr, envir = env)
      proc.time()[["elapsed"]] - t0
    }
  ))
}

run_noop <- function(fun, n) {
  for (i in seq_len(n)) {
    fun()
  }
  invisible(NULL)
}

run_rand <- function(fun, n, reps) {
  for (i in seq_len(reps)) {
    invisible(fun(n))
  }
  invisible(NULL)
}

run_fill <- function(fun, n, reps) {
  for (i in seq_len(reps)) {
    out <- numeric(n)
    invisible(fun(out, n))
  }
  invisible(NULL)
}

rtinycc_recipe <- tcc_ffi() |>
  tcc_source(rtinycc_code) |>
  tcc_bind(
    noop = list(args = list(), returns = "void"),
    fill_rand = list(args = list("numeric_array", "i32"), returns = "void"),
    rand_unif = list(
      args = list("i32"),
      returns = list(type = "numeric_array", length_arg = 1, free = TRUE)
    )
  )

generated_code <- Rtinycc:::generate_ffi_code(
  symbols = rtinycc_recipe$symbols,
  headers = rtinycc_recipe$headers,
  c_code = rtinycc_recipe$c_code,
  is_external = FALSE,
  structs = rtinycc_recipe$structs,
  unions = rtinycc_recipe$unions,
  enums = rtinycc_recipe$enums,
  globals = rtinycc_recipe$globals,
  container_of = rtinycc_recipe$container_of,
  field_addr = rtinycc_recipe$field_addr,
  struct_raw_access = rtinycc_recipe$struct_raw_access,
  introspect = rtinycc_recipe$introspect
)

compile_times <- data.frame(
  implementation = c("Rtinycc", "callme"),
  milliseconds = round(c(
    median_elapsed(build_rtinycc_module(), times = 2L),
    median_elapsed(build_and_dispose_callme_module(), times = 2L)
  ) * 1000, 1)
)

noop_bench <- with_benchmark_modules(function(rt_mod, cm_mod) {
  n_noop <- 1000L
  bench::mark(
    Rtinycc = run_noop(rt_mod$noop, n_noop),
    callme = run_noop(cm_mod$noop, n_noop),
    iterations = 20,
    check = TRUE,
    memory = TRUE,
    filter_gc = FALSE
  )
})

fill_bench_n4096 <- with_benchmark_modules(function(rt_mod, cm_mod) {
  bench::mark(
    Rtinycc = run_fill(rt_mod$fill_rand, 4096L, 100L),
    callme = run_fill(cm_mod$fill_rand, 4096L, 100L),
    iterations = 20,
    check = FALSE,
    memory = TRUE,
    filter_gc = FALSE
  )
})

rand_results <- with_benchmark_modules(function(rt_mod, cm_mod) {
  rand_bench_n1 <- bench::mark(
    Rtinycc = run_rand(rt_mod$rand_unif, 1L, 1000L),
    callme = run_rand(cm_mod$rand_unif, 1L, 1000L),
    iterations = 20,
    check = FALSE,
    memory = TRUE,
    filter_gc = FALSE
  )

  rand_bench_n4096 <- bench::mark(
    Rtinycc = run_rand(rt_mod$rand_unif, 4096L, 100L),
    callme = run_rand(cm_mod$rand_unif, 4096L, 100L),
    iterations = 20,
    check = FALSE,
    memory = TRUE,
    filter_gc = FALSE
  )

  list(rand_bench_n1 = rand_bench_n1, rand_bench_n4096 = rand_bench_n4096)
})

rand_bench_n1 <- rand_results$rand_bench_n1
rand_bench_n4096 <- rand_results$rand_bench_n4096

compile_times
cat(generated_code)
noop_bench
fill_bench_n4096
rand_bench_n1
rand_bench_n4096
```

The usual pattern is:

- `Rtinycc` compiles tiny modules much faster because it stays in-process
- a regular `.Call()` module has lower minimal call overhead
- the zero-copy `fill_rand()` case is the fairer vector benchmark for `Rtinycc`
- when `Rtinycc` returns a native buffer that must be copied into an R vector,
  that copy is measurable
- part of the runtime gap is also expected backend quality: `callme` is using
  the system compiler, while `Rtinycc` is using TinyCC

For performance-sensitive paths, the main mitigation is to move more work into
each call. One large array-oriented call is usually a much better fit than many
tiny scalar crossings. The full benchmark discussion is in the
`Compilation and Call Overhead` vignette.



### Variadic calls (e.g. `Rprintf` style)

Rtinycc supports two ways to bind variadic tails. The legacy approach uses
`varargs` as a typed prefix tail, while the bounded dynamic approach uses
`varargs_types` together with `varargs_min` and `varargs_max`. In the bounded
mode, wrappers are generated across the allowed arity and type combinations,
and runtime dispatch selects the matching wrapper from the scalar tail values
provided at call time.

```{r ffi-variadic-rprintf}
ffi_var <- tcc_ffi() |>
  tcc_header("#include <R_ext/Print.h>") |>
  tcc_source('
    #include <stdarg.h>

    int sum_fmt(int n, ...) {
      va_list ap;
      va_start(ap, n);
      int s = 0;
      for (int i = 0; i < n; i++) s += va_arg(ap, int);
      va_end(ap);
      Rprintf("sum_fmt(%d) = %d\\n", n, s);
      return s;
    }
  ') |>
  tcc_bind(
    Rprintf = list(
      args = list("cstring"),
      variadic = TRUE,
      varargs_types = list("i32"),
      varargs_min = 0L,
      varargs_max = 4L,
      returns = "void"
    ),
    sum_fmt = list(
      args = list("i32"),
      variadic = TRUE,
      varargs_types = list("i32"),
      varargs_min = 0L,
      varargs_max = 4L,
      returns = "i32"
    )
  ) |>
  tcc_compile()

ffi_var$Rprintf("Rprintf via bind: %d + %d = %d\n", 2L, 3L, 5L)
ffi_var$sum_fmt(0L)
ffi_var$sum_fmt(2L, 10L, 20L)
ffi_var$sum_fmt(4L, 1L, 2L, 3L, 4L)
```


### Linking external libraries

We can bind directly to symbols in shared libraries. Here we link against `libm`.

```{r ffi-link}
math <- tcc_ffi() |>
  tcc_library("m") |>
  tcc_bind(
    sqrt  = list(args = list("f64"), returns = "f64"),
    sin   = list(args = list("f64"), returns = "f64"),
    floor = list(args = list("f64"), returns = "f64")
  ) |>
  tcc_compile()

math$sqrt(16.0)
math$sin(pi / 2)
math$floor(3.7)



```
### Compiler options

Use `tcc_options()` to pass raw TinyCC options in the high-level FFI pipeline.
For low-level states, use `tcc_set_options()` directly.

```{r ffi-options}
ffi_opt_off <- tcc_ffi() |>
  tcc_options("-O0") |>
  tcc_source('
    int opt_macro() {
    #ifdef __OPTIMIZE__
      return 1;
    #else
      return 0;
    #endif
    }
  ') |>
  tcc_bind(opt_macro = list(args = list(), returns = "i32")) |>
  tcc_compile()

ffi_opt_on <- tcc_ffi() |>
  tcc_options(c("-Wall", "-O2")) |>
  tcc_source('
    int opt_macro() {
    #ifdef __OPTIMIZE__
      return 1;
    #else
      return 0;
    #endif
    }
  ') |>
  tcc_bind(opt_macro = list(args = list(), returns = "i32")) |>
  tcc_compile()

ffi_opt_off$opt_macro()
ffi_opt_on$opt_macro()
```
### Working with arrays

R vectors are passed to C with zero copy. Mutations in C are visible in R.

```{r ffi-arrays}
ffi <- tcc_ffi() |>
  tcc_source("
    #include <stdlib.h>
    #include <string.h>

    int64_t sum_array(int32_t* arr, int32_t n) {
      int64_t s = 0;
      for (int i = 0; i < n; i++) s += arr[i];
      return s;
    }

    void bump_first(int32_t* arr) { arr[0] += 10; }

    int32_t* dup_array(int32_t* arr, int32_t n) {
      int32_t* out = malloc(sizeof(int32_t) * n);
      memcpy(out, arr, sizeof(int32_t) * n);
      return out;
    }
  ") |>
  tcc_bind(
    sum_array  = list(args = list("integer_array", "i32"), returns = "i64"),
    bump_first = list(args = list("integer_array"), returns = "void"),
    dup_array  = list(
      args = list("integer_array", "i32"),
      returns = list(type = "integer_array", length_arg = 2, free = TRUE)
    )
  ) |>
  tcc_compile()

x <- as.integer(1:100) # to avoid ALTREP
.Internal(inspect(x))
ffi$sum_array(x, length(x))

# Zero-copy: C mutation reflects in R
ffi$bump_first(x)
x[1]

# Array return: copied into a new R vector, C buffer freed
y <- ffi$dup_array(x, length(x))
y[1]

.Internal(inspect(x))
```
## Advanced FFI features

### Structs and unions

Complex C types are supported declaratively. Use `tcc_struct()` to generate allocation and accessor helpers. Free instances when done.

```{r struct-example}
ffi <- tcc_ffi() |>
  tcc_source('
    #include <math.h>
    struct point { double x; double y; };
    double distance(struct point* a, struct point* b) {
      double dx = a->x - b->x, dy = a->y - b->y;
      return sqrt(dx * dx + dy * dy);
    }
  ') |>
  tcc_library("m") |>
  tcc_struct("point", accessors = c(x = "f64", y = "f64")) |>
  tcc_bind(distance = list(args = list("ptr", "ptr"), returns = "f64")) |>
  tcc_compile()

p1 <- ffi$struct_point_new()
ffi$struct_point_set_x(p1, 0.0)
ffi$struct_point_set_y(p1, 0.0)

p2 <- ffi$struct_point_new()
ffi$struct_point_set_x(p2, 3.0)
ffi$struct_point_set_y(p2, 4.0)

ffi$distance(p1, p2)

ffi$struct_point_free(p1)
ffi$struct_point_free(p2)
```

### Enums

Enums are exposed as helper functions that return integer constants.

```{r enum-example}
ffi <- tcc_ffi() |>
  tcc_source("enum color { RED = 0, GREEN = 1, BLUE = 2 };") |>
  tcc_enum("color", constants = c("RED", "GREEN", "BLUE")) |>
  tcc_compile()

ffi$enum_color_RED()
ffi$enum_color_BLUE()
```

### Bitfields

Bitfields are handled by TCC. Accessors read and write them like normal fields.

```{r bitfield-example}
ffi <- tcc_ffi() |>
  tcc_source("
    struct flags {
      unsigned int active : 1;
      unsigned int level  : 4;
    };
  ") |>
  tcc_struct("flags", accessors = c(active = "u8", level = "u8")) |>
  tcc_compile()

s <- ffi$struct_flags_new()
ffi$struct_flags_set_active(s, 1L)
ffi$struct_flags_set_level(s, 9L)
ffi$struct_flags_get_active(s)
ffi$struct_flags_get_level(s)
ffi$struct_flags_free(s)
```

### Global getters and setters

C globals can be exposed with explicit getter/setter helpers.

```{r globals-example}
ffi <- tcc_ffi() |>
  tcc_source("
    int counter = 7;
    double pi_approx = 3.14159;
  ") |>
  tcc_global("counter", "i32") |>
  tcc_global("pi_approx", "f64") |>
  tcc_compile()

ffi$global_counter_get()
ffi$global_pi_approx_get()
ffi$global_counter_set(42L)
ffi$global_counter_get()
```

### Callbacks

R functions can be registered as C function pointers via `tcc_callback()` and passed to compiled code. Specify a `callback:<signature>` argument in `tcc_bind()` so the trampoline is generated automatically. Call `tcc_callback_close()` when you want deterministic invalidation and earlier release of the preserved R function.

```{r callback-example}
cb <- tcc_callback(function(x) x * x, signature = "double (*)(double)")

code <- '
double apply_fn(double (*fn)(void* ctx, double), void* ctx, double x) {
  return fn(ctx, x);
}
'

ffi <- tcc_ffi() |>
  tcc_source(code) |>
  tcc_bind(
    apply_fn = list(
      args = list("callback:double(double)", "ptr", "f64"),
      returns = "f64"
    )
  ) |>
  tcc_compile()

ffi$apply_fn(cb, tcc_callback_ptr(cb), 7.0)
tcc_callback_close(cb)
```

### Callback errors

If a callback throws an R error, the trampoline catches it, emits a warning, and returns a type-appropriate sentinel instead of unwinding through C. In practice this means NA-like numeric or integer values, `NA` logical, `NA` string, or a null external pointer depending on the declared return type.

```{r callback-error}
cb_err <- tcc_callback(
  function(x) stop("boom"),
  signature = "double (*)(double)"
)

ffi_err <- tcc_ffi() |>
  tcc_source('
    double call_cb_err(double (*cb)(void* ctx, double), void* ctx, double x) {
      return cb(ctx, x);
    }
  ') |>
  tcc_bind(
    call_cb_err = list(
      args = list("callback:double(double)", "ptr", "f64"),
      returns = "f64"
    )
  ) |>
  tcc_compile()

warned <- FALSE
res <- withCallingHandlers(
  ffi_err$call_cb_err(cb_err, tcc_callback_ptr(cb_err), 1.0),
  warning = function(w) {
    warned <<- TRUE
    invokeRestart("muffleWarning")
  }
)
list(warned = warned, result = res)
tcc_callback_close(cb_err)
```

### Async callbacks

For thread-safe scheduling from worker threads, use
`callback_async:<signature>` in `tcc_bind()`. The async callback queue
is initialized automatically at package load.

When a bound function has any `callback_async:` argument, the generated
wrapper automatically runs your C function on a new thread while draining
callbacks on the main R thread. Your C code doesn't need to know about
draining at all — just call the callback as normal and the wrapper handles
the rest.

**Void return (fire-and-forget):** the callback is enqueued from any
thread and executed on the main R thread automatically — on Windows via
R's message pump, on Linux/macOS via R's event loop `addInputHandler`.

**Non-void return (synchronous):** the worker thread blocks until the
main R thread executes the callback and returns the real result.
Supported return types: integer variants (`int`, `int32_t`, `i8`, `i16`,
`u8`, `u16`), floating-point (`double`, `float`), `bool`/`logical`, and
pointer (`void*`, `T*`).

```{r callback-async}
# Fire-and-forget: void callback accumulated from 100 worker threads
hits <- 0L
cb_async <- tcc_callback(
  function(x) { hits <<- hits + x; NULL },
  signature = "void (*)(int)"
)

code_async <- '
struct task { void (*cb)(void* ctx, int); void* ctx; int value; };

#ifdef _WIN32
#include <windows.h>

static DWORD WINAPI worker(LPVOID data) {
  struct task* t = (struct task*) data;
  t->cb(t->ctx, t->value);
  return 0;
}

int spawn_async(void (*cb)(void* ctx, int), void* ctx, int value) {
  if (!cb || !ctx) return -1;
  struct task t;
  t.cb = cb;
  t.ctx = ctx;
  t.value = value;
  HANDLE th = CreateThread(NULL, 0, worker, &t, 0, NULL);
  if (!th) return -2;
  WaitForSingleObject(th, INFINITE);
  CloseHandle(th);
  return 0;
}
#else
#include <pthread.h>

static void* worker(void* data) {
  struct task* t = (struct task*) data;
  t->cb(t->ctx, t->value);
  return NULL;
}

int spawn_async(void (*cb)(void* ctx, int), void* ctx, int value) {
  if (!cb || !ctx) return -1;
  const int n = 100;
  struct task tasks[100];
  pthread_t th[100];
  for (int i = 0; i < n; i++) {
    tasks[i].cb = cb;
    tasks[i].ctx = ctx;
    tasks[i].value = value;
    if (pthread_create(&th[i], NULL, worker, &tasks[i]) != 0) {
      for (int j = 0; j < i; j++) pthread_join(th[j], NULL);
      return -2;
    }
  }
  for (int i = 0; i < n; i++) pthread_join(th[i], NULL);
  return 0;
}
#endif
'

ffi_async <- tcc_ffi() |>
  tcc_source(code_async)
if (.Platform$OS.type != "windows") {
  ffi_async <- tcc_library(ffi_async, "pthread")
}
ffi_async <- ffi_async |>
  tcc_bind(
    spawn_async = list(
      args = list("callback_async:void(int)", "ptr", "i32"),
      returns = "i32"
    )
  ) |>
  tcc_compile()

rc <- ffi_async$spawn_async(cb_async, tcc_callback_ptr(cb_async), 2L)
hits
tcc_callback_close(cb_async)
```

Non-void return works the same way — the generated wrapper handles the
drain loop transparently:

```{r callback-async-nonvoid}
cb_triple <- tcc_callback(
  function(x) x * 3L,
  signature = "int (*)(int)"
)

# Pure C: the worker calls the sync callback and returns its result.
# No drain logic needed — the generated wrapper handles it.
code_sync <- '
#ifdef _WIN32
#include <windows.h>

struct itask { int (*cb)(void*,int); void* ctx; int in; int out; };

static DWORD WINAPI iworker(LPVOID p) {
  struct itask* t = (struct itask*)p;
  t->out = t->cb(t->ctx, t->in);
  return 0;
}
int run_worker(int (*cb)(void*,int), void* ctx, int x) {
  struct itask t;
  t.cb = cb; t.ctx = ctx; t.in = x; t.out = -1;
  HANDLE th = CreateThread(NULL, 0, iworker, &t, 0, NULL);
  if (!th) return -1;
  WaitForSingleObject(th, INFINITE);
  CloseHandle(th);
  return t.out;
}
#else
#include <pthread.h>

struct itask { int (*cb)(void*,int); void* ctx; int in; int out; };

static void* iworker(void* p) {
  struct itask* t = (struct itask*)p;
  t->out = t->cb(t->ctx, t->in);
  return NULL;
}
int run_worker(int (*cb)(void*,int), void* ctx, int x) {
  struct itask t;
  t.cb = cb; t.ctx = ctx; t.in = x; t.out = -1;
  pthread_t th;
  if (pthread_create(&th, NULL, iworker, &t) != 0) return -1;
  pthread_join(th, NULL);
  return t.out;
}
#endif
'

ffi_sync <- tcc_ffi() |>
  tcc_source(code_sync)
if (.Platform$OS.type != "windows") {
  ffi_sync <- tcc_library(ffi_sync, "pthread")
}
ffi_sync <- ffi_sync |>
  tcc_bind(
    run_worker = list(args = list("callback_async:int(int)", "ptr", "i32"), returns = "i32")
  ) |>
  tcc_compile()

ffi_sync$run_worker(cb_triple, tcc_callback_ptr(cb_triple), 7L)  # 21
tcc_callback_close(cb_triple)
```

### SQLite: a complete example

This example ties together external library linking, callbacks, and pointer dereferencing. We open an in-memory SQLite database, execute queries, and collect rows through an R callback that reads `char**` arrays using `tcc_read_ptr` and `tcc_read_cstring`.

```{r ffi-sqlite, eval=TRUE}
ptr_size <- .Machine$sizeof.pointer

read_string_array <- function(ptr, n) {
  vapply(seq_len(n), function(i) {
    tcc_read_cstring(tcc_read_ptr(ptr, (i - 1L) * ptr_size))
  }, "")
}

cb <- tcc_callback(
  function(argc, argv, cols) {
    values <- read_string_array(argv, argc)
    names  <- read_string_array(cols, argc)
    cat(paste(names, values, sep = " = ", collapse = ", "), "\n")
    0L
  },
  signature = "int (*)(int, char **, char **)"
)

sqlite <- tcc_ffi() |>
  tcc_header("#include <sqlite3.h>") |>
  tcc_library("sqlite3") |>
  tcc_source('
    void* open_db() {
      sqlite3* db = NULL;
      sqlite3_open(":memory:", &db);
      return db;
    }
    int close_db(void* db) {
      return sqlite3_close((sqlite3*)db);
    }
  ') |>
  tcc_bind(
    open_db  = list(args = list(), returns = "ptr"),
    close_db = list(args = list("ptr"), returns = "i32"),
    sqlite3_libversion = list(args = list(), returns = "cstring"),
    sqlite3_exec = list(
      args = list("ptr", "cstring", "callback:int(int, char **, char **)", "ptr", "ptr"),
      returns = "i32"
    )
  ) |>
  tcc_compile()

sqlite$sqlite3_libversion()

db <- sqlite$open_db()
sqlite$sqlite3_exec(db, "CREATE TABLE t (id INTEGER, name TEXT);", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$sqlite3_exec(db, "INSERT INTO t VALUES (1, 'hello'), (2, 'world');", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$sqlite3_exec(db, "SELECT * FROM t;", cb, tcc_callback_ptr(cb), tcc_null_ptr())
sqlite$close_db(db)
tcc_callback_close(cb)
```

## Header parsing with treesitter.c

For header-driven bindings, we use `treesitter.c` to parse function signatures and generate binding specifications automatically. For struct, enum, and global helpers, `tcc_generate_bindings()` handles the code generation.

The default mapper is conservative for pointers: `char*` is treated as `ptr` because C does not guarantee NUL-terminated strings. If you know a parameter is a C string, provide a custom mapper that returns `cstring` for that type.

```{r ffi-treesitter}
header <- '
double sqrt(double x);
double sin(double x);
struct point { double x; double y; };
enum status { OK = 0, ERROR = 1 };
int global_counter;
'

tcc_treesitter_functions(header)
tcc_treesitter_structs(header)
tcc_treesitter_enums(header)
tcc_treesitter_globals(header)

# Bind parsed functions to libm
symbols <- tcc_treesitter_bindings(header)
math <- tcc_link("m", symbols = symbols)
math$sqrt(16.0)

# Generate struct/enum/global helpers
ffi <- tcc_ffi() |>
  tcc_source(header) |>
  tcc_generate_bindings(
    header,
    functions = FALSE, structs = TRUE,
    enums = TRUE, globals = TRUE
  ) |>
  tcc_compile()

ffi$struct_point_new()
ffi$enum_status_OK()
ffi$global_global_counter_get()
```

## io_uring Demo

`CSV` parser using [`io_uring`](https://en.wikipedia.org/wiki/Io_uring) on linux

```{r io_uring-demo, eval=TRUE}
if (Sys.info()[["sysname"]] == "Linux") {
  c_file <- system.file("c_examples", "io_uring_csv.c", package = "Rtinycc")

  n_rows <- 20000L
  n_cols <- 8L
  block_size <- 1024L * 1024L

  set.seed(42)
  tmp_csv <- tempfile("rtinycc_io_uring_readme_", fileext = ".csv")
  on.exit(unlink(tmp_csv), add = TRUE)

  mat <- matrix(runif(n_rows * n_cols), ncol = n_cols)
  df <- as.data.frame(mat)
  names(df) <- paste0("V", seq_len(n_cols))
  utils::write.table(df, file = tmp_csv, sep = ",", row.names = FALSE, col.names = TRUE, quote = FALSE)
  csv_size_mb <- as.double(file.info(tmp_csv)$size) / 1024^2
  message(sprintf("CSV size: %.2f MB", csv_size_mb))

  io_uring_src <- paste(readLines(c_file, warn = FALSE), collapse = "\n")

  ffi <- tcc_ffi() |>
    tcc_source(io_uring_src) |>
    tcc_bind(
      csv_table_read = list(
        args = list("cstring", "i32", "i32"),
        returns = "sexp"
      ),
      csv_table_io_uring = list(
        args = list("cstring", "i32", "i32"),
        returns = "sexp"
      )
    ) |>
    tcc_compile()

  baseline <- utils::read.table(tmp_csv, sep = ",", header = TRUE)
  c_tbl <- ffi$csv_table_read(tmp_csv, block_size, n_cols)
  uring_tbl <- ffi$csv_table_io_uring(tmp_csv, block_size, n_cols)
  vroom_tbl <- vroom::vroom(
    tmp_csv,
    delim = ",",
    altrep = FALSE,
    col_types = vroom::cols(.default = "d"),
    progress = FALSE,
    show_col_types = FALSE
  )

  stopifnot(
    identical(dim(c_tbl), dim(baseline)),
    identical(dim(uring_tbl), dim(baseline)),
    identical(dim(vroom_tbl), dim(baseline)),
    isTRUE(all.equal(c_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE)),
    isTRUE(all.equal(uring_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE)),
    isTRUE(all.equal(vroom_tbl, baseline, tolerance = 1e-8, check.attributes = FALSE))
  )

  timings <- bench::mark(
    read_table_df = {
      x <- utils::read.table(tmp_csv, sep = ",", header = TRUE)
      nrow(x)
    },
    vroom_df_altrep_false = {
      x <- vroom::vroom(
        tmp_csv,
        delim = ",",
        altrep = FALSE,
        col_types = vroom::cols(.default = "d"),
        progress = FALSE,
        show_col_types = FALSE
      )
      nrow(x)
    },
    vroom_df_altrep_false_mat = {
      x <- vroom::vroom(
        tmp_csv,
        delim = ",",
        altrep = FALSE,
        col_types = vroom::cols(.default = "d"),
        progress = FALSE,
        show_col_types = FALSE
      )
      x <- as.matrix(x)
      nrow(x)
    },
    c_read_df = {
      x <- ffi$csv_table_read(tmp_csv, block_size, n_cols)
      nrow(x)
    },
    io_uring_df = {
      x <- ffi$csv_table_io_uring(tmp_csv, block_size, n_cols)
      nrow(x)
    },
    iterations = 2,
    memory = TRUE
  )

  
  print(timings)
  
  plot(timings, type = "boxplot") + bench::scale_x_bench_time(base = NULL)
}
```

## Known limitations

### `_Complex` types

TCC does not support C99 `_Complex` types. Generated code works around this with `#define _Complex`, which suppresses the keyword. Apply the same workaround in your own `tcc_source()` code when headers pull in complex types.

### 64-bit integer precision

R represents `i64` and `u64` values as `double`, which loses precision beyond $2^{53}$. Values that differ only past that threshold become indistinguishable.

```{r limits-int64}
sprintf("2^53:     %.0f", 2^53)
sprintf("2^53 + 1: %.0f", 2^53 + 1)
identical(2^53, 2^53 + 1)
```

For exact 64-bit arithmetic, keep values in C-allocated storage and manipulate them through pointers.

### Nested structs

The accessor generator does not handle nested structs by value. Use pointer fields instead and reach inner structs with `tcc_field_addr()`.

```{r limits-nested}
ffi <- tcc_ffi() |>
  tcc_source('
    struct inner { int a; };
    struct outer { struct inner* in; };
  ') |>
  tcc_struct("inner", accessors = c(a = "i32")) |>
  tcc_struct("outer", accessors = c(`in` = "ptr")) |>
  tcc_field_addr("outer", "in") |>
  tcc_compile()

o <- ffi$struct_outer_new()
i <- ffi$struct_inner_new()
ffi$struct_inner_set_a(i, 42L)

# Write the inner pointer into the outer struct
ffi$struct_outer_in_addr(o) |> tcc_ptr_set(i)

# Read it back through indirection
ffi$struct_outer_in_addr(o) |>
  tcc_data_ptr() |>
  ffi$struct_inner_get_a()

ffi$struct_inner_free(i)
ffi$struct_outer_free(o)
```

### Array fields in structs

Array fields require the `list(type = ..., size = N, array = TRUE)` syntax in `tcc_struct()`, which generates element-wise accessors.

```{r limits-arrays}
ffi <- tcc_ffi() |>
  tcc_source('struct buf { unsigned char data[16]; };') |>
  tcc_struct("buf", accessors = list(
    data = list(type = "u8", size = 16, array = TRUE)
  )) |>
  tcc_compile()

b <- ffi$struct_buf_new()
ffi$struct_buf_set_data_elt(b, 0L, 0xCAL)
ffi$struct_buf_set_data_elt(b, 1L, 0xFEL)
ffi$struct_buf_get_data_elt(b, 0L)
ffi$struct_buf_get_data_elt(b, 1L)
ffi$struct_buf_free(b)
```

## Serialization and fork safety

Compiled FFI objects are fork-safe: `parallel::mclapply()` and other `fork()`-based parallelism work out of the box because TCC's compiled code lives in memory mappings that survive `fork()` via copy-on-write.

Serialization is also supported. Each `tcc_compiled` object stores its FFI recipe internally, so after `saveRDS()` / `readRDS()` (or `serialize()` / `unserialize()`), the first `$` access detects the dead TCC state pointer and recompiles transparently.

```{r serialize-example}
ffi <- tcc_ffi() |>
  tcc_source("int square(int x) { return x * x; }") |>
  tcc_bind(square = list(args = list("i32"), returns = "i32")) |>
  tcc_compile()

ffi$square(7L)

tmp <- tempfile(fileext = ".rds")
saveRDS(ffi, tmp)
ffi2 <- readRDS(tmp)
unlink(tmp)

# Auto-recompiles on first access
ffi2$square(7L)
```

For explicit control, use `tcc_recompile()`. Note that raw `tcc_state` objects and bare pointers from `tcc_malloc()` do not carry a recipe and remain dead after deserialization.

## License

GPL-3

## References

- [TinyCC](https://github.com/TinyCC/tinycc)
- [Bun's FFI](https://bun.com/docs/runtime/ffi)
- [CFFI](https://cffi.readthedocs.io/)
- [RSimpleFFI](https://github.com/sounkou-bioinfo/RSimpleFFI#readme)
- [CSlug](https://cslug.readthedocs.io/en/latest/)
