---
title: "Compilation and Call Overhead"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compilation and Call Overhead}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(Rtinycc)
has_callme <- requireNamespace("callme", quietly = TRUE)
```

This article measures two different costs:

- compilation latency for a tiny module
- call overhead once the code is already compiled

The comparison target is the [`callme`](https://cran.r-project.org/package=callme)
package, which builds ordinary `.Call()` entry points with `R CMD SHLIB`. That
gives us a useful baseline against a conventional R C API module.

The point is not that the two packages expose identical APIs. They do not.
Instead, the comparison asks a narrower question:

- how much compile-time latency does in-memory TinyCC avoid?
- what is the extra runtime cost of Rtinycc's generated wrapper layer?
- how much does an extra copy matter when Rtinycc has to convert a returned C
  buffer into an R vector?

## Two Minimal Cases

We use two tiny workloads:

- `noop()`: takes nothing, returns nothing
- `rand_unif(n)`: generates `n` random doubles

The `callme` version of `rand_unif()` allocates the final R vector directly with
the R C API.

The `Rtinycc` version returns a heap-allocated `double*`, and the generated
wrapper copies that buffer into a fresh R numeric vector before freeing the
original C allocation.

That makes the second case useful for isolating copy overhead.

```{r}
rtinycc_code <- '
  #include <R.h>
  #include <Rinternals.h>
  #include <Rmath.h>
  #include <stdlib.h>

  void noop(void) {}

  double* rand_unif(int n) {
    if (n < 0) {
      Rf_error("n must be non-negative");
    }
    if (n == 0) {
      return (double*) malloc(sizeof(double));
    }

    double *out = (double*) malloc(sizeof(double) * (size_t) n);
    if (!out) {
      Rf_error("malloc failed");
    }

    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      out[i] = unif_rand();
    }
    PutRNGstate();
    return out;
  }
'

callme_code <- '
  #include <R.h>
  #include <Rinternals.h>
  #include <Rmath.h>

  SEXP noop(void) {
    return R_NilValue;
  }

  SEXP rand_unif(SEXP n_) {
    int n = asInteger(n_);
    if (n < 0) {
      Rf_error("n must be non-negative");
    }

    SEXP out = PROTECT(allocVector(REALSXP, n));
    double *ptr = REAL(out);

    GetRNGstate();
    for (int i = 0; i < n; ++i) {
      ptr[i] = unif_rand();
    }
    PutRNGstate();

    UNPROTECT(1);
    return out;
  }
'

build_rtinycc_module <- function() {
  tcc_ffi() |>
    tcc_source(rtinycc_code) |>
    tcc_bind(
      noop = list(args = list(), returns = "void"),
      rand_unif = list(
        args = list("i32"),
        returns = list(type = "numeric_array", length_arg = 1, free = TRUE)
      )
    ) |>
    tcc_compile()
}

build_callme_module <- function() {
  callme::compile(callme_code, env = NULL, verbosity = 0)
}

median_elapsed <- function(expr, times = 3L) {
  expr <- substitute(expr)
  env <- parent.frame()
  stats::median(replicate(
    times,
    {
      gc()
      t0 <- proc.time()[["elapsed"]]
      eval(expr, envir = env)
      proc.time()[["elapsed"]] - t0
    }
  ))
}

run_noop <- function(fun, n) {
  for (i in seq_len(n)) {
    fun()
  }
  invisible(NULL)
}

run_rand <- function(fun, n, reps) {
  for (i in seq_len(reps)) {
    invisible(fun(n))
  }
  invisible(NULL)
}
```

## Availability

```{r}
has_callme
```

If this prints `FALSE`, the executable benchmarks below are skipped.

## Compilation Latency

This measures module build time, not call time.

```{r, eval = has_callme}
compile_times <- data.frame(
  implementation = c("Rtinycc", "callme"),
  seconds = c(
    median_elapsed(build_rtinycc_module(), times = 3L),
    median_elapsed(build_callme_module(), times = 3L)
  )
)

compile_times$milliseconds <- round(compile_times$seconds * 1000, 1)
compile_times
```

The expected pattern is:

- `Rtinycc` wins clearly on tiny compile latency because it stays in-process and
  does not shell out to `R CMD SHLIB`
- `callme` pays the ordinary shared-library toolchain cost

## `noop()` Call Overhead

This is the smallest useful call path. It approximates the lower bound on call
overhead above a plain `.Call()` entry point.

```{r, eval = has_callme}
rt_mod <- build_rtinycc_module()
cm_mod <- build_callme_module()

n_noop <- 50000L

noop_times <- data.frame(
  implementation = c("Rtinycc", "callme"),
  seconds = c(
    median_elapsed(run_noop(rt_mod$noop, n_noop), times = 5L),
    median_elapsed(run_noop(cm_mod$noop, n_noop), times = 5L)
  )
)

noop_times$ns_per_call <- round(noop_times$seconds / n_noop * 1e9, 1)
noop_times
```

Interpretation:

- the `callme` path is close to the cost of a conventional `.Call()` wrapper
- the `Rtinycc` path adds the generated wrapper layer and external-pointer call
  target
- the difference here is mostly boundary overhead, not useful computation

## `rand_unif(n)` And Copy Cost

Here the implementation work is still small, but the return path differs:

- `callme` fills the final R vector directly
- `Rtinycc` fills a native buffer, then the wrapper copies into a fresh R vector

We time both a tiny and a larger return size.

```{r, eval = has_callme}
rand_specs <- data.frame(
  n = c(1L, 4096L),
  reps = c(5000L, 300L)
)

rand_times <- do.call(
  rbind,
  lapply(seq_len(nrow(rand_specs)), function(i) {
    n <- rand_specs$n[[i]]
    reps <- rand_specs$reps[[i]]

    data.frame(
      implementation = c("Rtinycc", "callme"),
      n = n,
      reps = reps,
      seconds = c(
        median_elapsed(run_rand(rt_mod$rand_unif, n, reps), times = 5L),
        median_elapsed(run_rand(cm_mod$rand_unif, n, reps), times = 5L)
      )
    )
  })
)

rand_times$us_per_call <- round(rand_times$seconds / rand_times$reps * 1e6, 1)
rand_times
```

The usual pattern is:

- for `n = 1`, wrapper overhead and return-path mechanics dominate
- for larger `n`, the copy still matters, but more of the time is spent in the
  actual loop and RNG generation

## What These Numbers Mean

The benchmark gives a reasonable mental model:

- `Rtinycc` is optimized for low compilation latency and direct interactive use
- for very small scalar calls, a traditional `.Call()` entry point has lower
  overhead
- when `Rtinycc` must copy returned buffers into R vectors, that copy is real
  and measurable
- the main way to amortize the boundary cost is to do more work per call

So the package is usually strongest when:

- compile latency matters
- you want to bind plain C signatures quickly
- you batch work into array-oriented or coarse-grained calls

It is less ideal when:

- every microsecond of scalar-call overhead matters
- you can already afford and manage a regular shared-library toolchain
- you need a direct `.Call()` entry point that writes its final result straight
  into R-managed objects
